---
title: "Untitled"
output: html_document
date: "2026-07-24"
---
```{r include=FALSE}
# install required pacakges
install.packages(plyr)
install.packages(dplyr)
install.packages(mgcv)
install.packages(table1)
install.packages(tableone)
install.packages(qgcomp)
install.packages(qgcompint)
install.packages(lubridate)
install.packages(ggplot2)
install.packages(tidyr)
install.packages(tidyverse)
install.packages(sf)
install.packages(patchwork)
```


# run libraries
```{r include=FALSE}
library(plyr)
library(dplyr)
library(mgcv)
library(table1)
library(tableone)
library(qgcomp)
library(qgcompint)
library(lubridate)
library(ggplot2)
library(tidyr)
library(tidyverse)
library(sf)
library(patchwork)
```


save the data to a specific location

# Set working direcory
```{r}
wd = "C:/......"

```

# Import data

List of variables:
# year - monitoring year
# Gender- Male/Female
# Race4- Race in 4 categories
# edu4 -  Education level in 4 categories
# work2 - Occupation on 9/11
# exp911_imp2 - WTC exposure index (High/Very High=1/ Low/Intermediate=0)
# arr_9_11_13 - Arrival date (Arrived on 9/11-9/14 =1/ Arrived after 9/14 =0)
# horror - Witnessing horror (yes=1/ No=0)
# smoke2 -  Smoking status
# age - Age at monitoring visit
# non_na_count - Number of counted deficits
# pct_black - % Black population 
# pct_poverty - % under the poverty line
# pct_NoHS - % with no high school education
# MEDHH - Median household income 
# pm - PM2.5 (µg m-3)
# mean_temp - Mean temperature (° C)
# Mean_GVI_300 - Green View Index inside 300m buffer
# Mean_GVI_500 - Green View Index inside 500m buffer
# frailty.sum - Frailty index is calculated using 30 items that are collected during each monitoring visit, including deficits such as diabetes, fair/poor self-rated health, difficulty climbing several flights of stairs, abnormal heart examination, and self-reported mental health conditions. Each item is coded as a binary value where a response of Fair/Poor is considered as a “deficit” (and assigned a score of 1), and a response of Excellent/Very Good/Good is assigned a score of 0 (i.e., not a deficit). 
# PTSD - PTSD score based on the DSM-IV measurement 
# depression - Depression score based on PHQ9.

```{r}
data = read.csv(paste0(wd, '/Sim_Data.csv'))
head(data)
```

# Subset data for frailty analysis. The population for the frailty analysis is restricted to rows with at least 27 deficits. 
#Rows with less than 27 deficite are considered missing due to insufficient data to calculate the frailty index.  
```{r}
data_index= subset(data, non_na_count >=27)
```

# Remove outliers from the datasets 
# df_trimmed = The data for the frailty analysis
# df_trimmed2 = The data for the mental illness analyses
```{r}
#Function to remove lower and upper 0.5% in the exposure distributions to avoid bias due to extreme values. 
trim_df <- function(data, vars, p = 0.005) {
  # Loop over variables and compute quantiles
  for (v in vars) {
    lower <- quantile(data[[v]], p, na.rm = TRUE)
    upper <- quantile(data[[v]], 1 - p, na.rm = TRUE)
    
    # Keep only rows within range
    data <- data[data[[v]] >= lower & data[[v]] <= upper, ]
  }
  return(data)
}

# Apply to temperature and PM exposures
df_trimmed <- trim_df(data_index, vars = c('pm','mean_temp'))

df_trimmed <- df_trimmed %>%
  filter(!if_all(everything(), is.na))

df_trimmed2 <- trim_df(data, vars = c('pm','mean_temp'))

df_trimmed2 <- df_trimmed2 %>%
  filter(!if_all(everything(), is.na))
```

# Figure S4: outcome distributions desriptives 
# Frailty
```{r}
# Create dataframe
df <- data.frame(frailty = data_index$frailty.sum)

# Mean & SD
mu <- mean(df$frailty, na.rm = TRUE)
sigma <- sd(df$frailty, na.rm = TRUE)

f = ggplot(df, aes(x = frailty)) +
  geom_histogram(aes(y = ..density..),
                 bins = 30,
                 fill = "skyblue",
                 color = "white",
                 alpha = 0.9) +
  stat_function(fun = dnorm,
                args = list(mean = mu, sd = sigma),
                color = "darkred",
                linewidth = 1.2) +
  theme_minimal(base_size = 14) +
  labs(
    x = "Frailty Index",
    y = "Density"
  ) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    panel.grid.minor = element_blank()
  )
f
```

# PTSD
```{r}
# Create dataframe
df2 <- data.frame(ptsd_hist = data$PTSD)

# Mean & SD
mu <- mean(df2$ptsd_hist, na.rm = TRUE)
sigma <- sd(df2$ptsd_hist, na.rm = TRUE)

f1 = ggplot(df2, aes(x = ptsd_hist)) +
  geom_histogram(aes(y = ..density..),
                 bins = 30,
                 fill = "skyblue",
                 color = "white",
                 alpha = 0.9) +
  stat_function(fun = dnorm,
                args = list(mean = mu, sd = sigma),
                color = "darkred",
                linewidth = 1.2) +
  theme_minimal(base_size = 14) +
  labs(
    x = "PTSD Score",
    y = "Density"
  ) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    panel.grid.minor = element_blank()
  )
f1
```

# Depression
```{r}
# Create dataframe
df3 <- data.frame(dep_hist = data$depression)

# Mean & SD
mu <- mean(df3$dep_hist, na.rm = TRUE)
sigma <- sd(df3$dep_hist, na.rm = TRUE)

f2 = ggplot(df3, aes(x = dep_hist)) +
  geom_histogram(aes(y = ..density..),
                 bins = 30,
                 fill = "skyblue",
                 color = "white",
                 alpha = 0.9) +
  stat_function(fun = dnorm,
                args = list(mean = mu, sd = sigma),
                color = "darkred",
                linewidth = 1.2) +
  theme_minimal(base_size = 14) +
  labs(
    x = "Depression Score",
    y = "Density"
  ) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    panel.grid.minor = element_blank()
  )
f2
```

# combine plots into one
```{r}
f  <- f  +  theme(
  plot.title = element_text(size = 18),
  axis.title = element_text(size = 20),
  axis.text  = element_text(size = 18)
)
f1 <- f1 + theme(
  plot.title = element_text(size = 18),
  axis.title = element_text(size = 20),
  axis.text  = element_text(size = 18)
)
f2 <- f2 + theme(
  plot.title = element_text(size = 18),
  axis.title = element_text(size = 20),
  axis.text  = element_text(size = 18)
)

combined_plot <- f + f1 + f2 

combined_plot
```
