---
title: "Analysis code"
author: "Helena Krasnov"
date: "2026-05-01"
output: html_document
---

This code represents the original analytic workflow used in the study. Since the original data cannot be shared due to data-use and confidentiality restrictions, we created a simulated dataset to demonstrate code reproducibility and analytic procedures. As a result, the outputs generated from the simulated data differ from the results presented in the manuscript.

The simulation data is located in the zip file attached to this code

```{r include=FALSE}
# install required pacakges
install.packages(plyr)
install.packages(dplyr)
install.packages(mgcv)
install.packages(table1)
install.packages(tableone)
install.packages(qgcomp)
install.packages(qgcompint)
install.packages(lubridate)
install.packages(ggplot2)
install.packages(tidyr)
install.packages(tidyverse)
install.packages(sf)
install.packages(patchwork)
```


# run libraries
```{r include=FALSE}
library(plyr)
library(dplyr)
library(mgcv)
library(table1)
library(tableone)
library(qgcomp)
library(qgcompint)
library(lubridate)
library(ggplot2)
library(tidyr)
library(tidyverse)
library(sf)
library(patchwork)
```


save the data to a specific location

# Set working direcory
```{r}
wd = "C:/......"

```

# Import data

List of variables:
# year - monitoring year
# Gender- Male/Female
# Race4- Race in 4 categories
# edu4 -  Education level in 4 categories
# work2 - Occupation on 9/11
# exp911_imp2 - WTC exposure index (High/Very High=1/ Low/Intermediate=0)
# arr_9_11_13 - Arrival date (Arrived on 9/11-9/14 =1/ Arrived after 9/14 =0)
# horror - Witnessing horror (yes=1/ No=0)
# smoke2 -  Smoking status
# age - Age at monitoring visit
# non_na_count - Number of counted deficits
# pct_black - % Black population 
# pct_poverty - % under the poverty line
# pct_NoHS - % with no high school education
# MEDHH - Median household income 
# pm - PM2.5 (µg m-3)
# mean_temp - Mean temperature (° C)
# Mean_GVI_300 - Green View Index inside 300m buffer
# Mean_GVI_500 - Green View Index inside 500m buffer
# frailty.sum - Frailty index is calculated using 30 items that are collected during each monitoring visit, including deficits such as diabetes, fair/poor self-rated health, difficulty climbing several flights of stairs, abnormal heart examination, and self-reported mental health conditions. Each item is coded as a binary value where a response of Fair/Poor is considered as a “deficit” (and assigned a score of 1), and a response of Excellent/Very Good/Good is assigned a score of 0 (i.e., not a deficit). 
# PTSD - PTSD score based on the DSM-IV measurement 
# depression - Depression score based on PHQ9.

```{r}
data = read.csv(paste0(wd, '/Sim_Data.csv'))
head(data)
```

# Descriptive prep function: used to create Table 1
```{r}
# change decimal point in table
my.render.cont <- function(x) {
    with(stats.default(x), 
         c("",
          "Mean (SD)" = sprintf("%s (%s)",
                                round_pad(MEAN, 2),
                                round_pad(SD, 2))))
}

my.render.cat <- function(x) { 
  sub('.', '.', c("", sapply(stats.default(x), 
  function(y) with(y, sprintf("%d (%0.2f %%)", FREQ, PCT)))), fixed = TRUE) 
}

# 1. Define the missing value renderer to match your style
my.render.miss <- function(x, ...) {
  with(stats.default(x), 
       c("", "Missing" = sprintf("%d (%0.2f %%)", NMISS, PCTMISS)))
}
```

## Table 1. Population characteristics at cohort entry (N = 18,734), 2003-2023
```{r}
m2 <-data %>% group_by(RunningID) %>% filter(row_number()==1)

m2$outcome = 'Descriptive Statistics'

table1(~ age + as.factor(Race4) + as.factor(edu4) + as.factor(Gender) + 
           as.factor(smoke2) + as.factor(work2) + MEDHH + pct_black + 
           pct_poverty + pct_NoHS | outcome, 
       data = m2,
       overall = FALSE,
       render.continuous = my.render.cont, 
       render.categorical = my.render.cat,
       render.missing = render.missing.default)
```

# Table 2. Summary statistics of annual exposures and modifiers
```{r}
# Create mean continuous variables 
tab <- data %>%
    group_by(RunningID) %>%
    summarise(
        pm = mean(pm, na.rm = TRUE),
        mean_temp = mean(mean_temp, na.rm = TRUE),
        Mean_GVI_300 = mean(Mean_GVI_300, na.rm = TRUE),
        # categorical variables just take first (they don't change anyway)
        exp911_imp2 = first(exp911_imp2),
        arr_9_11_13 = first(arr_9_11_13),
        horror = first(horror)
    )

# Define variables
listVars <- c('pm','mean_temp','Mean_GVI_300','exp911_imp2','arr_9_11_13','horror')
catVars  <- c('exp911_imp2','arr_9_11_13','horror')

# Create Table 2
table2 <- CreateTableOne(vars = listVars,
                         data = tab,
                         factorVars = catVars,
                         includeNA = TRUE)

print(table2)
```

# Subset data for frailty analysis. The population for the frailty analysis is restricted to rows with at least 27 deficits. 
#Rows with less than 27 deficite are considered missing due to insufficient data to calculate the frailty index.  
```{r}
data_index= subset(data, non_na_count >=27)
```

# Remove outliers from the datasets 
# df_trimmed = The data for the frailty analysis
# df_trimmed2 = The data for the mental illness analyses
```{r}
#Function to remove lower and upper 0.5% in the exposure distributions to avoid bias due to extreme values. 
trim_df <- function(data, vars, p = 0.005) {
  # Loop over variables and compute quantiles
  for (v in vars) {
    lower <- quantile(data[[v]], p, na.rm = TRUE)
    upper <- quantile(data[[v]], 1 - p, na.rm = TRUE)
    
    # Keep only rows within range
    data <- data[data[[v]] >= lower & data[[v]] <= upper, ]
  }
  return(data)
}

# Apply to temperature and PM exposures
df_trimmed <- trim_df(data_index, vars = c('pm','mean_temp'))

df_trimmed <- df_trimmed %>%
  filter(!if_all(everything(), is.na))

df_trimmed2 <- trim_df(data, vars = c('pm','mean_temp'))

df_trimmed2 <- df_trimmed2 %>%
  filter(!if_all(everything(), is.na))
```

# Compute IQR for the main exposures 
```{r}
var <- c("pm", "mean_temp", "Mean_GVI_300")

iqrs <- sapply(var, function(var) IQR(data[[var]], na.rm = TRUE))
```

# Table 3. The association between environmental exposures and Frailty, PTSD, and Depression scores.
# Model 1: Frialty analysis
```{r}
frail_mod <- bam(frailty.sum~ pm +mean_temp+ Mean_GVI_300 +age+s(year,bs='ps') + as.factor(Gender) + as.factor(Race4) + as.factor(edu4)+ pct_black + MEDHH + pct_poverty+pct_NoHS+as.factor(work2)+ as.factor(smoke2)+s(RunningID,bs='re'),data=df_trimmed, family = quasipoisson(link = log),offset = log(non_na_count))

    pm=summary(frail_mod)$p.table
    pm<-as.data.frame(pm,keep.rownames = TRUE)[] # Format as datatable
    table_pm=pm[2:4,]
    
    # get upper and lower bound
    table_pm$lower = table_pm$Estimate - qnorm(0.975) * table_pm$`Std. Error`
    table_pm$upper = table_pm$Estimate + qnorm(0.975) * table_pm$`Std. Error`
    table_pm = table_pm[,c(1,5:6) ]

    # PM row
    table_pm$Estimate_exp <- (exp(table_pm$Estimate*iqrs)-1)*100
    table_pm$lower_exp <- (exp(table_pm$lower*iqrs)-1)*100
    table_pm$upper_exp <- (exp(table_pm$upper*iqrs)-1)*100
    
  # Round values 
  table_pm$Estimate_exp <- round(table_pm$Estimate_exp, 3)
  table_pm$lower_exp <- round(table_pm$lower_exp, 3)
  table_pm$upper_exp <- round(table_pm$upper_exp, 3)
  table_pm$complete <- paste(table_pm$Estimate_exp, "(", table_pm$lower_exp, ";", table_pm$upper_exp, ")")
  table_pm[, 7, drop = FALSE]
```

# Table 3. The association between environmental exposures and Frailty, PTSD, and Depression scores.
# Model 2: PTSD analysis
```{r}
#Outcome model
ptsd_mod <- bam(as.numeric(PTSD)~ pm +mean_temp+  Mean_GVI_300  +age+s(year,bs='ps') + as.factor(Gender) + as.factor(Race4) + as.factor(edu4)+ pct_black + MEDHH + pct_poverty+pct_NoHS+as.factor(work2)+ as.factor(smoke2)+s(RunningID,bs='re'),family = quasipoisson(link = log), data=df_trimmed2)

    pm=summary(ptsd_mod)$p.table
    pm<-as.data.frame(pm,keep.rownames = TRUE)[] # Format as datatable
    table_pm=pm[2:4,]
    
    # get upper and lower bound
    table_pm$lower = table_pm$Estimate - qnorm(0.975) * table_pm$`Std. Error`
    table_pm$upper = table_pm$Estimate + qnorm(0.975) * table_pm$`Std. Error`
    table_pm = table_pm[,c(1,5:6) ]

    # PM row
    table_pm$Estimate_exp <- (exp(table_pm$Estimate*iqrs)-1)*100
    table_pm$lower_exp <- (exp(table_pm$lower*iqrs)-1)*100
    table_pm$upper_exp <- (exp(table_pm$upper*iqrs)-1)*100
    
  # Round values 
  table_pm$Estimate_exp <- round(table_pm$Estimate_exp, 3)
  table_pm$lower_exp <- round(table_pm$lower_exp, 3)
  table_pm$upper_exp <- round(table_pm$upper_exp, 3)
  table_pm$complete <- paste(table_pm$Estimate_exp, "(", table_pm$lower_exp, ";", table_pm$upper_exp, ")")
  table_pm[, 7, drop = FALSE]

```

# Table 3. The association between environmental exposures and Frailty, PTSD, and Depression scores.
# Model 3 - Depression
```{r}
PHQ9_mod <- bam(as.numeric(depression)~ pm +mean_temp+  Mean_GVI_300  +age+s(year,bs='ps') + as.factor(Gender) + as.factor(Race4) + as.factor(edu4)+ pct_black + MEDHH + pct_poverty+pct_NoHS+as.factor(work2)+ as.factor(smoke2)+s(RunningID,bs='re'),family = quasipoisson(link = log),data=df_trimmed2) 

    pm=summary(PHQ9_mod)$p.table
    pm<-as.data.frame(pm,keep.rownames = TRUE)[] # Format as datatable
    table_pm=pm[2:4,]
    
    table_pm$lower = table_pm$Estimate - qnorm(0.975) * table_pm$`Std. Error`
    table_pm$upper = table_pm$Estimate + qnorm(0.975) * table_pm$`Std. Error`
    table_pm = table_pm[,c(1,5:6) ]

    # PM row
    table_pm$Estimate_exp <- (exp(table_pm$Estimate*iqrs)-1)*100
    table_pm$lower_exp <- (exp(table_pm$lower*iqrs)-1)*100
    table_pm$upper_exp <- (exp(table_pm$upper*iqrs)-1)*100
    
  # Round values 
  table_pm$Estimate_exp <- round(table_pm$Estimate_exp, 3)
  table_pm$lower_exp <- round(table_pm$lower_exp, 3)
  table_pm$upper_exp <- round(table_pm$upper_exp, 3)
  table_pm$complete <- paste(table_pm$Estimate_exp, "(", table_pm$lower_exp, ";", table_pm$upper_exp, ")")
  table_pm[, 7, drop = FALSE]

```

# Mixture analysis # Table 3 and Figure 1
```{r}
mixture =  c("pm", "mean_temp", 'Mean_GVI_300')
```

# Frailty table 3
```{r}
# Outcome model for frailty
mod_ipw = qgcomp.noboot(frailty.sum ~  
                      pm +  Mean_GVI_300 +mean_temp +age+as.factor(year) + as.factor(work2)+ as.factor(Race4) + as.factor(Gender)+ as.factor(edu4)+ pct_black + MEDHH +pct_poverty+pct_NoHS+ as.factor(smoke2)+offset(log(non_na_count)),id = df_trimmed$ExtractID,bayes = TRUE, expnms = mixture, data=df_trimmed ,q=10, family=quasipoisson(link = "log"))

summary(mod_ipw)
```

# Frailty Figure 1
```{r}
# create plot Figure 1
# --- Extract weights ---
pos_df <- as.data.frame(mod_ipw$pos.weights)
neg_df <- as.data.frame(mod_ipw$neg.weights)

pos_df <- tibble(variable = rownames(pos_df), weight = pos_df[[1]], direction = "Positive")
neg_df <- tibble(variable = rownames(neg_df), weight = -neg_df[[1]], direction = "Negative")  # flip sign

# Combine and clean
weights_df <- bind_rows(pos_df, neg_df) %>%
    filter(weight != 0)

# --- Clean labels (with math formatting) ---
weights_df$variable_label <- recode(weights_df$variable,
                                    "pm" = "PM[2.5]~'(μg/m'^3*')'",
                                    "mean_temp" = "'Mean Temperature (°C)'",
                                    "Mean_GVI_300" = "'Green View Index (300 m)'"
)

# --- Order so PM2.5 appears at TOP after coord_flip() ---
weights_df$variable_label <- factor(
    weights_df$variable_label,
    levels = rev(c(  # reverse order for top-down display
        "PM[2.5]~'(μg/m'^3*')'",
        "'Mean Temperature (°C)'",
        "'Green View Index (300 m)'"
    ))
)

# --- Plot ---
p <- ggplot(weights_df, aes(x = variable_label, y = weight, fill = direction)) +
    geom_bar(stat = "identity", color = "black", width = 0.75, alpha = 0.9) +
    coord_flip() +
    geom_hline(yintercept = 0, color = "gray40", linewidth = 0.6) +
    scale_fill_manual(values = c("Positive" = "#4C566A", "Negative" = "#A7ADBA")) +
    theme_minimal(base_size = 14) +
    labs(
        title = "Frailty Index",
        x = NULL,
        y = "Weight",
        #fill = "Direction"
    ) +
    theme(
        plot.title = element_text(face = "bold", size = 15, hjust = 0.5),
        panel.grid.major.y = element_blank(),
        legend.position = "none"
    ) +
    scale_x_discrete(labels = function(x) parse(text = x))
```

# Mixture analysis: PTSD # Table 3
```{r}
# --- Run outcome model ---
mix_ptsd = qgcomp.noboot(
    as.numeric(PTSD) ~  
        pm + Mean_GVI_300 + mean_temp + age + as.factor(year) +
        as.factor(Race4) + as.factor(edu4) + pct_black + MEDHH +
        pct_poverty + pct_NoHS + as.factor(work2) + as.factor(smoke2)+ as.factor(Gender),
    id = df_trimmed2$ExtractID,
    bayes = TRUE,
    expnms = mixture,
    data = df_trimmed2,
    q = 10,
    family = quasipoisson(link = "log"))

summary(mix_ptsd)
```

# PTSD Figure 1
```{r}
# Figure 1
# --- Extract weights ---
pos_df <- as.data.frame(mix_ptsd$pos.weights)
neg_df <- as.data.frame(mix_ptsd$neg.weights)

pos_df <- tibble(variable = rownames(pos_df), weight = pos_df[[1]], direction = "Positive")
neg_df <- tibble(variable = rownames(neg_df), weight = -neg_df[[1]], direction = "Negative")  # flip sign

# Combine and clean
weights_df <- bind_rows(pos_df, neg_df) %>%
    filter(weight != 0)

# --- Clean labels (with math formatting) ---
weights_df$variable_label <- recode(weights_df$variable,
                                    "pm" = "PM[2.5]~'(μg/m'^3*')'",
                                    "mean_temp" = "'Mean Temperature (°C)'",
                                    "Mean_GVI_300" = "'Green View Index (300 m)'"
)

# --- Order so PM2.5 appears at TOP after coord_flip() ---
weights_df$variable_label <- factor(
    weights_df$variable_label,
    levels = rev(c(  # reverse order for top-down display
        "PM[2.5]~'(μg/m'^3*')'",
        "'Mean Temperature (°C)'",
        "'Green View Index (300 m)'"
    ))
)

# --- Plot ---
p1 <- ggplot(weights_df, aes(x = variable_label, y = weight, fill = direction)) +
    geom_bar(stat = "identity", color = "black", width = 0.75, alpha = 0.9) +
    coord_flip() +
    geom_hline(yintercept = 0, color = "gray40", linewidth = 0.6) +
    scale_fill_manual(values = c("Positive" = "#4C566A", "Negative" = "#A7ADBA")) +
    theme_minimal(base_size = 14) +
    labs(
        title = "PTSD",
        x = NULL,
        y = "Weight",
        fill = "Direction"
    ) +
    theme(
        plot.title = element_text(face = "bold", size = 15, hjust = 0.5),
        panel.grid.major.y = element_blank(),
        legend.position = "bottom"
    ) +
    scale_x_discrete(labels = function(x) parse(text = x))

```


# Mixture analysis: Depression # Table 3

```{r}
#Outcome model
mix_dep = qgcomp.noboot(as.numeric(depression) ~  
                      pm +  Mean_GVI_300 +mean_temp +age+as.factor(year) +  as.factor(Race4) + as.factor(edu4)+ pct_black + MEDHH + as.factor(Gender)+ pct_poverty+pct_NoHS+as.factor(work2)+as.factor(smoke2),id = df_trimmed2$ExtractID,
                      bayes = TRUE, expnms = mixture, data=df_trimmed2, q=10, family=quasipoisson(link = "log"))

summary(mix_dep)
```

# Depression Figure 1
```{r}

# --- Extract weights ---
pos_df <- as.data.frame(mix_dep$pos.weights)
neg_df <- as.data.frame(mix_dep$neg.weights)

pos_df <- tibble(variable = rownames(pos_df), weight = pos_df[[1]], direction = "Positive")
neg_df <- tibble(variable = rownames(neg_df), weight = -neg_df[[1]], direction = "Negative")  # flip sign

# Combine and clean
weights_df <- bind_rows(pos_df, neg_df) %>%
    filter(weight != 0)

# --- Clean labels (with math formatting) ---
weights_df$variable_label <- recode(weights_df$variable,
                                    "pm" = "PM[2.5]~'(μg/m'^3*')'",
                                    "mean_temp" = "'Mean Temperature (°C)'",
                                    "Mean_GVI_300" = "'Green View Index (300 m)'"
)

# --- Order so PM2.5 appears at TOP after coord_flip() ---
weights_df$variable_label <- factor(
    weights_df$variable_label,
    levels = rev(c(  # reverse order for top-down display
        "PM[2.5]~'(μg/m'^3*')'",
        "'Mean Temperature (°C)'",
        "'Green View Index (300 m)'"
    ))
)

# --- Plot ---
p2 <- ggplot(weights_df, aes(x = variable_label, y = weight, fill = direction)) +
    geom_bar(stat = "identity", color = "black", width = 0.75, alpha = 0.9) +
    coord_flip() +
    geom_hline(yintercept = 0, color = "gray40", linewidth = 0.6) +
    scale_fill_manual(values = c("Positive" = "#4C566A", "Negative" = "#A7ADBA")) +
    theme_minimal(base_size = 14) +
    labs(
        title = "Depression",
        x = NULL,
        y = "Weight",
        fill = "Direction"
    ) +
    theme(
        plot.title = element_text(face = "bold", size = 15, hjust = 0.5),
        panel.grid.major.y = element_blank(),
        legend.position = "none"
    ) +
    scale_x_discrete(labels = function(x) parse(text = x))
```

# merge plots -Figure 1
```{r}
library(gridExtra)

grid.arrange(p,p1,p2,ncol=2)

```

# to calculate the % change in outcome 
```{r}
(exp(Estimate)-1)*100
(exp(Lower CI)-1)*100
(exp(Upper CI)-1)*100
```


# Table 4. Effect modification is in code Table 4


#######################################################################################
#######################################secondary analysis
#######################################################################################

# Table S2: Secondary analysis: PTSD and Depression (Yes/No) .
# PTSD, binomial, with cutoff of 44
```{r}
# Recode into a binary variable using 44 as cutoff for PTSD
df_trimmed2$ptsd = ifelse(df_trimmed2$PTSD<=44,0,1)

ptsd_mod2 <- bam(ptsd~ pm +mean_temp+  Mean_GVI_300  +age+s(year,bs='ps') + as.factor(Gender) + as.factor(Race4) + as.factor(edu4)+ pct_black + MEDHH + pct_poverty+pct_NoHS+as.factor(work2)+ as.factor(smoke2)+s(RunningID,bs='re'),data=df_trimmed2, family = binomial())

    pm=summary(ptsd_mod2)$p.table
    pm<-as.data.frame(pm,keep.rownames = TRUE)[] # Format as datatable
    table_pm=pm[2:4,]
    
    # get upper and lower bound
    table_pm$lower = table_pm$Estimate - qnorm(0.975) * table_pm$`Std. Error`
    table_pm$upper = table_pm$Estimate + qnorm(0.975) * table_pm$`Std. Error`
    table_pm = table_pm[,c(1,5:6) ]

    # PM row
    table_pm$Estimate_exp <- exp(table_pm$Estimate*iqrs)
    table_pm$lower_exp <- exp(table_pm$lower*iqrs)
    table_pm$upper_exp <- exp(table_pm$upper*iqrs)
    
  # Round values 
  table_pm$Estimate_exp <- round(table_pm$Estimate_exp, 3)
  table_pm$lower_exp <- round(table_pm$lower_exp, 3)
  table_pm$upper_exp <- round(table_pm$upper_exp, 3)
  table_pm$complete <- paste(table_pm$Estimate_exp, "(", table_pm$lower_exp, ";", table_pm$upper_exp, ")")
  table_pm[, 7, drop = FALSE]
```

# Depression, binomial, with cutoff of 10
```{r}
# Recode into a binary variable
df_trimmed2$depp = ifelse(df_trimmed2$depression<=9,0,1)

PHQ9_mod2 <- bam(depp~ pm +mean_temp+  Mean_GVI_300  +age+s(year,bs='ps') + as.factor(Gender) + as.factor(Race4) + as.factor(edu4)+ pct_black + MEDHH + pct_poverty+pct_NoHS+as.factor(work2)+ as.factor(smoke2)+s(RunningID,bs='re'),data=df_trimmed2, family = binomial())

    pm=summary(PHQ9_mod2)$p.table
    pm<-as.data.frame(pm,keep.rownames = TRUE)[] # Format as datatable
    table_pm=pm[2:4,]
    
    # get upper and lower bound
    table_pm$lower = table_pm$Estimate - qnorm(0.975) * table_pm$`Std. Error`
    table_pm$upper = table_pm$Estimate + qnorm(0.975) * table_pm$`Std. Error`
    table_pm = table_pm[,c(1,5:6) ]

    # PM row
    table_pm$Estimate_exp <- exp(table_pm$Estimate*iqrs)
    table_pm$lower_exp <- exp(table_pm$lower*iqrs)
    table_pm$upper_exp <- exp(table_pm$upper*iqrs)
    
  # Round values 
  table_pm$Estimate_exp <- round(table_pm$Estimate_exp, 3)
  table_pm$lower_exp <- round(table_pm$lower_exp, 3)
  table_pm$upper_exp <- round(table_pm$upper_exp, 3)
  table_pm$complete <- paste(table_pm$Estimate_exp, "(", table_pm$lower_exp, ";", table_pm$upper_exp, ")")
  table_pm[, 7, drop = FALSE]
```

#Table S2. Mixture analysis - PTSD, binomial, with cutoff of 44
```{r}
mix_ptsd2 = qgcomp.noboot(
    as.numeric(ptsd) ~  
        pm + Mean_GVI_300 + mean_temp + age + as.factor(year) +
        as.factor(Race4) + as.factor(edu4) + pct_black + MEDHH + as.factor(Gender)+
        pct_poverty + pct_NoHS + as.factor(work2) + as.factor(smoke2),
    id = df_trimmed2$ExtractID,
    bayes = TRUE,
    expnms = mixture,
    data = df_trimmed2,
    q = 10,
    family = binomial()
)

summary(mix_ptsd2)
```

#Table S2. Mixture analysis: Depression, binomial, with cutoff of 10
```{r}
mix_dep2 = qgcomp.noboot(depp ~  
                      pm +  Mean_GVI_300 +mean_temp +age+as.factor(year) +  as.factor(Race4) + as.factor(edu4)+ pct_black + MEDHH + as.factor(Gender)+ pct_poverty+pct_NoHS+as.factor(work2)+as.factor(smoke2),id = df_trimmed2$ExtractID,
                      bayes = TRUE, expnms = mixture, data=df_trimmed2, q=10, family=binomial())

summary(mix_dep2)
```

#Table S3. The association between environmental exposures and Frailty, PTSD, and Depression scores using Green View Index with 500m buffer
```{r}
var <- c("pm", "mean_temp", "Mean_GVI_500")

iqrs <- sapply(var, function(var) IQR(data[[var]], na.rm = TRUE))
```

# Associations with frailty
```{r}
pm_500 <-bam(frailty.sum~ pm +mean_temp+ Mean_GVI_500 +age+s(year,bs='ps') + as.factor(Gender) + as.factor(Race4) + as.factor(edu4)+ pct_black + MEDHH + pct_poverty+pct_NoHS+as.factor(work2)+ as.factor(smoke2)+s(RunningID,bs='re'),data=df_trimmed, family = quasipoisson(link = log),offset = log(non_na_count))

    pm=summary(pm_500)$p.table
    pm<-as.data.frame(pm,keep.rownames = TRUE)[] # Format as datatable
    table_pm=pm[2:4,]
    
    # get upper and lower bound
    table_pm$lower = table_pm$Estimate - qnorm(0.975) * table_pm$`Std. Error`
    table_pm$upper = table_pm$Estimate + qnorm(0.975) * table_pm$`Std. Error`
    table_pm = table_pm[,c(1,5:6) ]

    # PM row
    table_pm$Estimate_exp <- (exp(table_pm$Estimate*iqrs)-1)*100
    table_pm$lower_exp <- (exp(table_pm$lower*iqrs)-1)*100
    table_pm$upper_exp <- (exp(table_pm$upper*iqrs)-1)*100
    
  # Round values 
  table_pm$Estimate_exp <- round(table_pm$Estimate_exp, 3)
  table_pm$lower_exp <- round(table_pm$lower_exp, 3)
  table_pm$upper_exp <- round(table_pm$upper_exp, 3)
  table_pm$complete <- paste(table_pm$Estimate_exp, "(", table_pm$lower_exp, ";", table_pm$upper_exp, ")")
  table_pm[, 7, drop = FALSE]
```


# Associations with PTSD 
```{r}
ptsd_sd <- bam(as.numeric(PTSD)~ pm +mean_temp+  Mean_GVI_500  +age+s(year,bs='ps') + as.factor(Gender) + as.factor(Race4) + as.factor(edu4)+ pct_black + MEDHH + pct_poverty+pct_NoHS+as.factor(work2)+ as.factor(smoke2)+s(RunningID,bs='re'),family= quasipoisson(link = log),data=df_trimmed2)

    pm=summary(ptsd_sd)$p.table
    pm<-as.data.frame(pm,keep.rownames = TRUE)[] # Format as datatable
    table_pm=pm[2:4,]
    # get upper and lower bound
    table_pm$lower = table_pm$Estimate - qnorm(0.975) * table_pm$`Std. Error`
    table_pm$upper = table_pm$Estimate + qnorm(0.975) * table_pm$`Std. Error`
    table_pm = table_pm[,c(1,5:6) ]

    # PM row
    table_pm$Estimate_exp <- (exp(table_pm$Estimate*iqrs)-1)*100
    table_pm$lower_exp <- (exp(table_pm$lower*iqrs)-1)*100
    table_pm$upper_exp <- (exp(table_pm$upper*iqrs)-1)*100
    
  # Round values 
  table_pm$Estimate_exp <- round(table_pm$Estimate_exp, 3)
  table_pm$lower_exp <- round(table_pm$lower_exp, 3)
  table_pm$upper_exp <- round(table_pm$upper_exp, 3)
  table_pm$complete <- paste(table_pm$Estimate_exp, "(", table_pm$lower_exp, ";", table_pm$upper_exp, ")")
  table_pm[, 7, drop = FALSE]

```

# Association with Depression 
```{r}
PHQ9_sd <- bam(as.numeric(depression)~ pm +mean_temp+  Mean_GVI_500  +age+s(year,bs='ps') + as.factor(Gender) + as.factor(Race4) + as.factor(edu4)+ pct_black + MEDHH + pct_poverty+pct_NoHS+as.factor(work2)+ as.factor(smoke2)+s(RunningID,bs='re'),family= quasipoisson(link = log),data=df_trimmed2) 

    pm=summary(PHQ9_sd)$p.table
    pm<-as.data.frame(pm,keep.rownames = TRUE)[] # Format as datatable
    table_pm=pm[2:4,]
    
    # get upper and lower bound
    table_pm$lower = table_pm$Estimate - qnorm(0.975) * table_pm$`Std. Error`
    table_pm$upper = table_pm$Estimate + qnorm(0.975) * table_pm$`Std. Error`
    table_pm = table_pm[,c(1,5:6) ]

    # PM row
    table_pm$Estimate_exp <- (exp(table_pm$Estimate*iqrs)-1)*100
    table_pm$lower_exp <- (exp(table_pm$lower*iqrs)-1)*100
    table_pm$upper_exp <- (exp(table_pm$upper*iqrs)-1)*100
    
  # Round values 
  table_pm$Estimate_exp <- round(table_pm$Estimate_exp, 3)
  table_pm$lower_exp <- round(table_pm$lower_exp, 3)
  table_pm$upper_exp <- round(table_pm$upper_exp, 3)
  table_pm$complete <- paste(table_pm$Estimate_exp, "(", table_pm$lower_exp, ";", table_pm$upper_exp, ")")
  table_pm[, 7, drop = FALSE]
  
```

# Mixture models
```{r}
mixture2 =  c("pm", "mean_temp", 'Mean_GVI_500')

fi_500 = qgcomp.noboot(frailty.sum ~  
                      pm +  Mean_GVI_500 +mean_temp +age+as.factor(year) + as.factor(work2)+ as.factor(Race4) + as.factor(edu4)+ pct_black + MEDHH +pct_poverty+pct_NoHS+as.factor(Gender)+ as.factor(smoke2)+offset(log(non_na_count)),id = df_trimmed$ExtractID,bayes = TRUE, expnms = mixture2, data=df_trimmed ,q=10, family=quasipoisson(link = "log"))

summary(fi_500)

ptsd_500 = qgcomp.noboot(
    as.numeric(PTSD) ~  
        pm + Mean_GVI_500 + mean_temp + age + as.factor(year) +
        as.factor(Race4) + as.factor(edu4) + pct_black + MEDHH +as.factor(Gender)+
        pct_poverty + pct_NoHS + as.factor(work2) + as.factor(smoke2),
    id = df_trimmed2$ExtractID,
    bayes = TRUE,
    expnms = mixture2,
    data = df_trimmed2,
    q = 10,
    family=quasipoisson(link = "log")
)
summary(ptsd_500)

dep_500 = qgcomp.noboot(as.numeric(depression) ~  
                      pm +  Mean_GVI_500 +mean_temp +age+as.factor(year) +  as.factor(Race4) + as.factor(edu4)+ pct_black + MEDHH +  pct_poverty+pct_NoHS+as.factor(work2)+as.factor(Gender)+as.factor(smoke2),id = df_trimmed2$ExtractID,
                      bayes = TRUE, expnms = mixture2, data=df_trimmed2, q=10, family=quasipoisson(link = "log"))

summary(dep_500)
```

####################################################################
########################## Sup material ############################

# Figure S1A. Distribution of the responders' monitoring visits
#Data 'Figure1A.csv' is an aggregate summary of visit counts and frquancies in the cohort. 
```{r}
plot_FigureS1A = read.csv(paste0(wd, '/Figure1A.csv'))

ggplot(plot_FigureS1A, aes(x = factor(total_visits))) +
  geom_col(aes(y = count), fill = "skyblue", color = "grey") +
  geom_text(aes(y = count, label = label), vjust = -0.5, size = 3) +
  # cumulative line on secondary axis
  geom_line(aes(y = cumulative_pct / 100 * max(count), group = 1), 
            color = "darkred", linewidth = 1) +
  geom_point(aes(y = cumulative_pct / 100 * max(count)), 
             color = "darkred", size = 2) +
  geom_text(aes(y = cumulative_pct / 100 * max(count), label = cum_label), 
            vjust = -1.4, size = 2.5, color = "black") +
  scale_y_continuous(
    name = "Number of responders",
    sec.axis = sec_axis(~ . / max(plot_FigureS1A$count) * 100, 
                        name = "Cumulative %")
  ) +
  labs(
    title = "Distribution of study visits per person",
    x = "Visits Number"
  ) +
  theme_minimal() + annotate("text", x = Inf, y = Inf, label = "A", 
           hjust = 1.5, vjust = 2.5, fontface = "bold", size = 6)
```

# # Figure S1B showing the distribution of the first year of enrollmen
```{r}
plot_FigureS1B = read.csv(paste0(wd,'/Figure1B.csv'))

ggplot(plot_FigureS1B, aes(x = factor(year), y = row_count)) +
  geom_col(fill = "skyblue", color = "grey") + 
  geom_text(aes(label = paste0(round(100 * row_count / sum(row_count), 1), "%")),
            vjust = -0.3, size = 3) +
  labs(
    title = "Distribution of the first year of enrollment",
    x = NULL,
    y = "Number of responders"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)
  ) +
  annotate("text", x = Inf, y = Inf, label = "B", 
           hjust = 1.5, vjust = 2.5, fontface = "bold", size = 6)
```

#Figure S2.Flowchart of the study population. PPT

#Figure S3. Distribution of the responders' (person years) locations
```{r}
plot_FigureS3 = st_read(paste0(wd,'/FigureS3.shp'))

my_breaks <- seq(0, max(plot_FigureS3$pnt_cnt, na.rm = TRUE), by = 10)

library(ggplot2)

p <- ggplot(data = plot_FigureS3) +
  # 1. Map the fill to your variable
  geom_sf(aes(fill = pnt_cnt), colour = NA) + 
  
  # 2. Use the mako scale
  scale_fill_viridis_c(
    option = "mako", 
    begin = 0.4, 
    direction = -1,
    guide = guide_colorbar(
      title.position = "top",
      barwidth = unit(0.5, "cm"),
      barheight = unit(4, "cm")
    )
  ) +
  
  # 3. Apply theme and fix legend
  theme_void() + 
  labs(fill = "Person Years") +
  theme(
    legend.position = "right", # Changed from "none" to "right"
    legend.title = element_text(size = 10, face = "bold"),
    plot.background = element_rect(colour = "grey80", fill = NA, linewidth = 1),
    plot.margin = margin(10, 10, 10, 10) # Adds a little breathing room
  )

print(p)
```

# Figure S4: outcome distributions desriptives 
# Frailty
```{r}
# Create dataframe
df <- data.frame(frailty = data_index$frailty.sum)

# Mean & SD
mu <- mean(df$frailty, na.rm = TRUE)
sigma <- sd(df$frailty, na.rm = TRUE)

f = ggplot(df, aes(x = frailty)) +
  geom_histogram(aes(y = ..density..),
                 bins = 30,
                 fill = "skyblue",
                 color = "white",
                 alpha = 0.9) +
  stat_function(fun = dnorm,
                args = list(mean = mu, sd = sigma),
                color = "darkred",
                linewidth = 1.2) +
  theme_minimal(base_size = 14) +
  labs(
    x = "Frailty Index",
    y = "Density"
  ) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    panel.grid.minor = element_blank()
  )
f
```

# PTSD
```{r}
# Create dataframe
df2 <- data.frame(ptsd_hist = data$PTSD)

# Mean & SD
mu <- mean(df2$ptsd_hist, na.rm = TRUE)
sigma <- sd(df2$ptsd_hist, na.rm = TRUE)

f1 = ggplot(df2, aes(x = ptsd_hist)) +
  geom_histogram(aes(y = ..density..),
                 bins = 30,
                 fill = "skyblue",
                 color = "white",
                 alpha = 0.9) +
  stat_function(fun = dnorm,
                args = list(mean = mu, sd = sigma),
                color = "darkred",
                linewidth = 1.2) +
  theme_minimal(base_size = 14) +
  labs(
    x = "PTSD Score",
    y = "Density"
  ) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    panel.grid.minor = element_blank()
  )
f1
```

# Depression
```{r}
# Create dataframe
df3 <- data.frame(dep_hist = data$depression)

# Mean & SD
mu <- mean(df3$dep_hist, na.rm = TRUE)
sigma <- sd(df3$dep_hist, na.rm = TRUE)

f2 = ggplot(df3, aes(x = dep_hist)) +
  geom_histogram(aes(y = ..density..),
                 bins = 30,
                 fill = "skyblue",
                 color = "white",
                 alpha = 0.9) +
  stat_function(fun = dnorm,
                args = list(mean = mu, sd = sigma),
                color = "darkred",
                linewidth = 1.2) +
  theme_minimal(base_size = 14) +
  labs(
    x = "Depression Score",
    y = "Density"
  ) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    panel.grid.minor = element_blank()
  )
f2
```

# combine plots into one
```{r}
f  <- f  +  theme(
  plot.title = element_text(size = 18),
  axis.title = element_text(size = 20),
  axis.text  = element_text(size = 18)
)
f1 <- f1 + theme(
  plot.title = element_text(size = 18),
  axis.title = element_text(size = 20),
  axis.text  = element_text(size = 18)
)
f2 <- f2 + theme(
  plot.title = element_text(size = 18),
  axis.title = element_text(size = 20),
  axis.text  = element_text(size = 18)
)

combined_plot <- f + f1 + f2 

combined_plot
```

#Table S1. Number of deficits (person years) between 2003-2023
```{r}
TableS1 = read.csv(paste0(wd,'/deficits.csv'))

TableS1
```

#Table S4. The association between the exposures and Frailty: Sensitivity analysis incorporating the probability of having at least 27 items for the index calculation 
# Since 21% of responders had missing FI values, we added IPW and repeated the model as a sensitivity analysis. 
```{r}
#(1) Run logistic model to predict the probability of inclusion in our frailty analysis

# mark whether each row meets the “≥27 non-missing values”
data <- data %>%
  mutate(has27 = as.integer(non_na_count >= 27))

# model the flagged variable
prob_mod <- glm(
  has27 ~ age + as.factor(Gender) + as.factor(Race4) +
    as.factor(edu4) + pct_black + MEDHH + pct_poverty + pct_NoHS +
    as.factor(work2) + as.factor(smoke2),
  data = data,
  family = binomial(link = "logit")
)

# Predicted probability of having ≥27 non-NA values
prob_has27 <- predict(prob_mod, data, type = "response")
data$ipw_27 <- 1 / prob_has27

# take only rows that had more than 27 variables
data_ipw= subset(data, non_na_count >=27)

# remove outliers
data_ipw2 <- trim_df(data_ipw, vars = c('pm','mean_temp'))
data_ipw2 <- data_ipw2 %>%
  filter(!if_all(everything(), is.na))

```

```{r}
# Run the outcome model incorporating the IPW weights
pm_mean <-bam(frailty.sum~ pm +mean_temp+ Mean_GVI_300 +age+s(year,bs='ps') + as.factor(Gender) + as.factor(Race4) + as.factor(edu4)+ pct_black + MEDHH + pct_poverty+pct_NoHS+as.factor(work2)+ as.factor(smoke2)+s(RunningID,bs='re'),data=data_ipw2, family = quasipoisson(link = log),offset = log(non_na_count),weights = ipw_27)

    pm=summary(pm_mean)$p.table
    pm<-as.data.frame(pm,keep.rownames = TRUE)[] # Format as datatable
    table_pm=pm[2:4,]
    
    # get upper and lower bound
    table_pm$lower = table_pm$Estimate - qnorm(0.975) * table_pm$`Std. Error`
    table_pm$upper = table_pm$Estimate + qnorm(0.975) * table_pm$`Std. Error`
    table_pm = table_pm[,c(1,5:6) ]

    # PM row
    table_pm$Estimate_exp <- (exp(table_pm$Estimate*iqrs)-1)*100
    table_pm$lower_exp <- (exp(table_pm$lower*iqrs)-1)*100
    table_pm$upper_exp <- (exp(table_pm$upper*iqrs)-1)*100
    
  # Round values 
  table_pm$Estimate_exp <- round(table_pm$Estimate_exp, 3)
  table_pm$lower_exp <- round(table_pm$lower_exp, 3)
  table_pm$upper_exp <- round(table_pm$upper_exp, 3)
  table_pm$complete <- paste(table_pm$Estimate_exp, "(", table_pm$lower_exp, ";", table_pm$upper_exp, ")")
  table_pm[, 7, drop = FALSE]
```

# mixture with IPW
```{r}
# Outcome model for frailty (qgcomp with IPW)
mod_ipw = qgcomp.noboot(frailty.sum ~  
                      pm +  Mean_GVI_300 +mean_temp +age+as.factor(year) + as.factor(work2)+ as.factor(Race4) + as.factor(edu4)+ pct_black + MEDHH +pct_poverty+pct_NoHS+as.factor(Gender)+ as.factor(smoke2)+offset(log(non_na_count)),id = data_ipw2$RunningID,bayes = TRUE, expnms = mixture, data=data_ipw2 ,q=10, family=quasipoisson(link = "log"),weights = ipw_27)

summary(mod_ipw)
```
 
# Table S5. Sensitivity analysis restricting addresses only from 2012
#Create subsets for the farilty and PTSD/depression analyses
```{r}
subset_frail = subset(df_trimmed,year>=2012)

subset_ptsd = subset(df_trimmed2,year>=2012)
```

#Model for frailty
```{r}
frail_sub <- bam(frailty.sum~ pm +mean_temp+ Mean_GVI_300 +age+s(year,bs='ps') + as.factor(Gender) + as.factor(Race4) + as.factor(edu4)+ pct_black + MEDHH + pct_poverty+pct_NoHS+as.factor(work2)+ as.factor(smoke2)+s(RunningID,bs='re'),data=subset_frail, family = quasipoisson(link = log),offset = log(non_na_count))

    pm=summary(frail_sub)$p.table
    pm<-as.data.frame(pm,keep.rownames = TRUE)[] # Format as datatable
    table_pm=pm[2:4,]
    
    # get upper and lower bound
    table_pm$lower = table_pm$Estimate - qnorm(0.975) * table_pm$`Std. Error`
    table_pm$upper = table_pm$Estimate + qnorm(0.975) * table_pm$`Std. Error`
    table_pm = table_pm[,c(1,5:6) ]

    # PM row
    table_pm$Estimate_exp <- (exp(table_pm$Estimate*iqrs)-1)*100
    table_pm$lower_exp <- (exp(table_pm$lower*iqrs)-1)*100
    table_pm$upper_exp <- (exp(table_pm$upper*iqrs)-1)*100
    
  # Round values 
  table_pm$Estimate_exp <- round(table_pm$Estimate_exp, 3)
  table_pm$lower_exp <- round(table_pm$lower_exp, 3)
  table_pm$upper_exp <- round(table_pm$upper_exp, 3)
  table_pm$complete <- paste(table_pm$Estimate_exp, "(", table_pm$lower_exp, ";", table_pm$upper_exp, ")")
  table_pm[, 7, drop = FALSE]
```

#Model for PTSD
```{r}
#Outcome model
ptsd_sub <- bam(as.numeric(PTSD)~ pm +mean_temp+  Mean_GVI_300  +age+s(year,bs='ps') + as.factor(Gender) + as.factor(Race4) + as.factor(edu4)+ pct_black + MEDHH + pct_poverty+pct_NoHS+as.factor(work2)+ as.factor(smoke2)+s(RunningID,bs='re'),family = quasipoisson(link = log), data=subset_ptsd)

    pm=summary(ptsd_sub)$p.table
    pm<-as.data.frame(pm,keep.rownames = TRUE)[] # Format as datatable
    table_pm=pm[2:4,]
    
    # get upper and lower bound
    table_pm$lower = table_pm$Estimate - qnorm(0.975) * table_pm$`Std. Error`
    table_pm$upper = table_pm$Estimate + qnorm(0.975) * table_pm$`Std. Error`
    table_pm = table_pm[,c(1,5:6) ]

    # PM row
    table_pm$Estimate_exp <- (exp(table_pm$Estimate*iqrs)-1)*100
    table_pm$lower_exp <- (exp(table_pm$lower*iqrs)-1)*100
    table_pm$upper_exp <- (exp(table_pm$upper*iqrs)-1)*100
    
  # Round values 
  table_pm$Estimate_exp <- round(table_pm$Estimate_exp, 3)
  table_pm$lower_exp <- round(table_pm$lower_exp, 3)
  table_pm$upper_exp <- round(table_pm$upper_exp, 3)
  table_pm$complete <- paste(table_pm$Estimate_exp, "(", table_pm$lower_exp, ";", table_pm$upper_exp, ")")
  table_pm[, 7, drop = FALSE]

```
#Model for depression
```{r}
PHQ9_sub <- bam(as.numeric(depression)~ pm +mean_temp+  Mean_GVI_300  +age+s(year,bs='ps') + as.factor(Gender) + as.factor(Race4) + as.factor(edu4)+ pct_black + MEDHH + pct_poverty+pct_NoHS+as.factor(work2)+ as.factor(smoke2)+s(RunningID,bs='re'),family = quasipoisson(link = log),data=subset_ptsd) 

    pm=summary(PHQ9_sub)$p.table
    pm<-as.data.frame(pm,keep.rownames = TRUE)[] # Format as datatable
    table_pm=pm[2:4,]
    
    table_pm$lower = table_pm$Estimate - qnorm(0.975) * table_pm$`Std. Error`
    table_pm$upper = table_pm$Estimate + qnorm(0.975) * table_pm$`Std. Error`
    table_pm = table_pm[,c(1,5:6) ]

    # PM row
    table_pm$Estimate_exp <- (exp(table_pm$Estimate*iqrs)-1)*100
    table_pm$lower_exp <- (exp(table_pm$lower*iqrs)-1)*100
    table_pm$upper_exp <- (exp(table_pm$upper*iqrs)-1)*100
    
  # Round values 
  table_pm$Estimate_exp <- round(table_pm$Estimate_exp, 3)
  table_pm$lower_exp <- round(table_pm$lower_exp, 3)
  table_pm$upper_exp <- round(table_pm$upper_exp, 3)
  table_pm$complete <- paste(table_pm$Estimate_exp, "(", table_pm$lower_exp, ";", table_pm$upper_exp, ")")
  table_pm[, 7, drop = FALSE]
```

#Mixture models
```{r}
fi_500 = qgcomp.noboot(frailty.sum ~  
                      pm +  Mean_GVI_300 +mean_temp +age+as.factor(year) + as.factor(work2)+ as.factor(Race4) + as.factor(edu4)+ pct_black + MEDHH +pct_poverty+pct_NoHS+as.factor(Gender)+ as.factor(smoke2)+offset(log(non_na_count)),id = subset_frail$ExtractID,bayes = TRUE, expnms = mixture, data=subset_frail ,q=10, family=quasipoisson(link = "log"))

summary(fi_500)

ptsd_500 = qgcomp.noboot(
    as.numeric(PTSD) ~  
        pm + Mean_GVI_300 + mean_temp + age + as.factor(year) +
        as.factor(Race4) + as.factor(edu4) + pct_black + MEDHH +as.factor(Gender)+
        pct_poverty + pct_NoHS + as.factor(work2) + as.factor(smoke2),
    id = subset_ptsd$ExtractID,
    bayes = TRUE,
    expnms = mixture,
    data = subset_ptsd,
    q = 10,
    family=quasipoisson(link = "log")
)
summary(ptsd_500)

dep_500 = qgcomp.noboot(as.numeric(depression) ~  
                      pm +  Mean_GVI_300 +mean_temp +age+as.factor(year) +  as.factor(Race4) + as.factor(edu4)+ pct_black + MEDHH +  pct_poverty+pct_NoHS+as.factor(work2)+as.factor(Gender)+as.factor(smoke2),id = subset_ptsd$ExtractID,
                      bayes = TRUE, expnms = mixture, data=subset_ptsd, q=10, family=quasipoisson(link = "log"))

summary(dep_500)

```

