---
title: "Analysis code"
author: "Helena Krasnov"
date: "2026-05-01"
output: html_document
---

```{r include=FALSE}
# install required pacakges
install.packages(plyr)
install.packages(dplyr)
install.packages(mgcv)
install.packages(table1)
install.packages(tableone)
install.packages(qgcomp)
install.packages(qgcompint)
install.packages(lubridate)
install.packages(ggplot2)
install.packages(tidyr)
install.packages(tidyverse)
install.packages(sf)
install.packages(patchwork)
```


# run libraries
```{r include=FALSE}
library(plyr)
library(dplyr)
library(mgcv)
library(table1)
library(tableone)
library(qgcomp)
library(qgcompint)
library(lubridate)
library(ggplot2)
library(tidyr)
library(tidyverse)
library(sf)
library(patchwork)
```


save the data to a specific location

# Set working direcory
```{r}
wd = "C:/......"

```

# Import data

List of variables:
# year - monitoring year
# Gender- Male/Female
# Race4- Race in 4 categories
# edu4 -  Education level in 4 categories
# work2 - Occupation on 9/11
# exp911_imp2 - WTC exposure index (High/Very High=1/ Low/Intermediate=0)
# arr_9_11_13 - Arrival date (Arrived on 9/11-9/14 =1/ Arrived after 9/14 =0)
# horror - Witnessing horror (yes=1/ No=0)
# smoke2 -  Smoking status
# age - Age at monitoring visit
# non_na_count - Number of counted deficits
# pct_black - % Black population 
# pct_poverty - % under the poverty line
# pct_NoHS - % with no high school education
# MEDHH - Median household income 
# pm - PM2.5 (µg m-3)
# mean_temp - Mean temperature (° C)
# Mean_GVI_300 - Green View Index inside 300m buffer
# Mean_GVI_500 - Green View Index inside 500m buffer
# frailty.sum - Frailty index is calculated using 30 items that are collected during each monitoring visit, including deficits such as diabetes, fair/poor self-rated health, difficulty climbing several flights of stairs, abnormal heart examination, and self-reported mental health conditions. Each item is coded as a binary value where a response of Fair/Poor is considered as a “deficit” (and assigned a score of 1), and a response of Excellent/Very Good/Good is assigned a score of 0 (i.e., not a deficit). 
# PTSD - PTSD score based on the DSM-IV measurement 
# depression - Depression score based on PHQ9.

```{r}
data = read.csv(paste0(wd, '/Sim_Data.csv'))
head(data)
```

# Subset data for frailty analysis. The population for the frailty analysis is restricted to rows with at least 27 deficits. 
#Rows with less than 27 deficite are considered missing due to insufficient data to calculate the frailty index.  
```{r}
data_index= subset(data, non_na_count >=27)
```

# Remove outliers from the datasets 
# df_trimmed = The data for the frailty analysis
# df_trimmed2 = The data for the mental illness analyses
```{r}
#Function to remove lower and upper 0.5% in the exposure distributions to avoid bias due to extreme values. 
trim_df <- function(data, vars, p = 0.005) {
  # Loop over variables and compute quantiles
  for (v in vars) {
    lower <- quantile(data[[v]], p, na.rm = TRUE)
    upper <- quantile(data[[v]], 1 - p, na.rm = TRUE)
    
    # Keep only rows within range
    data <- data[data[[v]] >= lower & data[[v]] <= upper, ]
  }
  return(data)
}

# Apply to temperature and PM exposures
df_trimmed <- trim_df(data_index, vars = c('pm','mean_temp'))

df_trimmed <- df_trimmed %>%
  filter(!if_all(everything(), is.na))

df_trimmed2 <- trim_df(data, vars = c('pm','mean_temp'))

df_trimmed2 <- df_trimmed2 %>%
  filter(!if_all(everything(), is.na))
```

# Compute IQR for the main exposures 
```{r}
var <- c("pm", "mean_temp", "Mean_GVI_300")

iqrs <- sapply(var, function(var) IQR(data[[var]], na.rm = TRUE))
```

# Table 3. The association between environmental exposures and Frailty, PTSD, and Depression scores.
# Model 1: Frialty analysis
```{r}
frail_mod <- bam(frailty.sum~ pm +mean_temp+ Mean_GVI_300 +age+s(year,bs='ps') + as.factor(Gender) + as.factor(Race4) + as.factor(edu4)+ pct_black + MEDHH + pct_poverty+pct_NoHS+as.factor(work2)+ as.factor(smoke2)+s(RunningID,bs='re'),data=df_trimmed, family = quasipoisson(link = log),offset = log(non_na_count))

    pm=summary(frail_mod)$p.table
    pm<-as.data.frame(pm,keep.rownames = TRUE)[] # Format as datatable
    table_pm=pm[2:4,]
    
    # get upper and lower bound
    table_pm$lower = table_pm$Estimate - qnorm(0.975) * table_pm$`Std. Error`
    table_pm$upper = table_pm$Estimate + qnorm(0.975) * table_pm$`Std. Error`
    table_pm = table_pm[,c(1,5:6) ]

    # PM row
    table_pm$Estimate_exp <- (exp(table_pm$Estimate*iqrs)-1)*100
    table_pm$lower_exp <- (exp(table_pm$lower*iqrs)-1)*100
    table_pm$upper_exp <- (exp(table_pm$upper*iqrs)-1)*100
    
  # Round values 
  table_pm$Estimate_exp <- round(table_pm$Estimate_exp, 3)
  table_pm$lower_exp <- round(table_pm$lower_exp, 3)
  table_pm$upper_exp <- round(table_pm$upper_exp, 3)
  table_pm$complete <- paste(table_pm$Estimate_exp, "(", table_pm$lower_exp, ";", table_pm$upper_exp, ")")
  table_pm[, 7, drop = FALSE]
```

# Table 3. The association between environmental exposures and Frailty, PTSD, and Depression scores.
# Model 2: PTSD analysis
```{r}
#Outcome model
ptsd_mod <- bam(as.numeric(PTSD)~ pm +mean_temp+  Mean_GVI_300  +age+s(year,bs='ps') + as.factor(Gender) + as.factor(Race4) + as.factor(edu4)+ pct_black + MEDHH + pct_poverty+pct_NoHS+as.factor(work2)+ as.factor(smoke2)+s(RunningID,bs='re'),family = quasipoisson(link = log), data=df_trimmed2)

    pm=summary(ptsd_mod)$p.table
    pm<-as.data.frame(pm,keep.rownames = TRUE)[] # Format as datatable
    table_pm=pm[2:4,]
    
    # get upper and lower bound
    table_pm$lower = table_pm$Estimate - qnorm(0.975) * table_pm$`Std. Error`
    table_pm$upper = table_pm$Estimate + qnorm(0.975) * table_pm$`Std. Error`
    table_pm = table_pm[,c(1,5:6) ]

    # PM row
    table_pm$Estimate_exp <- (exp(table_pm$Estimate*iqrs)-1)*100
    table_pm$lower_exp <- (exp(table_pm$lower*iqrs)-1)*100
    table_pm$upper_exp <- (exp(table_pm$upper*iqrs)-1)*100
    
  # Round values 
  table_pm$Estimate_exp <- round(table_pm$Estimate_exp, 3)
  table_pm$lower_exp <- round(table_pm$lower_exp, 3)
  table_pm$upper_exp <- round(table_pm$upper_exp, 3)
  table_pm$complete <- paste(table_pm$Estimate_exp, "(", table_pm$lower_exp, ";", table_pm$upper_exp, ")")
  table_pm[, 7, drop = FALSE]

```

# Table 3. The association between environmental exposures and Frailty, PTSD, and Depression scores.
# Model 3 - Depression
```{r}
PHQ9_mod <- bam(as.numeric(depression)~ pm +mean_temp+  Mean_GVI_300  +age+s(year,bs='ps') + as.factor(Gender) + as.factor(Race4) + as.factor(edu4)+ pct_black + MEDHH + pct_poverty+pct_NoHS+as.factor(work2)+ as.factor(smoke2)+s(RunningID,bs='re'),family = quasipoisson(link = log),data=df_trimmed2) 

    pm=summary(PHQ9_mod)$p.table
    pm<-as.data.frame(pm,keep.rownames = TRUE)[] # Format as datatable
    table_pm=pm[2:4,]
    
    table_pm$lower = table_pm$Estimate - qnorm(0.975) * table_pm$`Std. Error`
    table_pm$upper = table_pm$Estimate + qnorm(0.975) * table_pm$`Std. Error`
    table_pm = table_pm[,c(1,5:6) ]

    # PM row
    table_pm$Estimate_exp <- (exp(table_pm$Estimate*iqrs)-1)*100
    table_pm$lower_exp <- (exp(table_pm$lower*iqrs)-1)*100
    table_pm$upper_exp <- (exp(table_pm$upper*iqrs)-1)*100
    
  # Round values 
  table_pm$Estimate_exp <- round(table_pm$Estimate_exp, 3)
  table_pm$lower_exp <- round(table_pm$lower_exp, 3)
  table_pm$upper_exp <- round(table_pm$upper_exp, 3)
  table_pm$complete <- paste(table_pm$Estimate_exp, "(", table_pm$lower_exp, ";", table_pm$upper_exp, ")")
  table_pm[, 7, drop = FALSE]

```

# Mixture analysis # Table 3 and Figure 1
```{r}
mixture =  c("pm", "mean_temp", 'Mean_GVI_300')
```

# Frailty table 3
```{r}
# Outcome model for frailty
mod_ipw = qgcomp.noboot(frailty.sum ~  
                      pm +  Mean_GVI_300 +mean_temp +age+as.factor(year) + as.factor(work2)+ as.factor(Race4) + as.factor(Gender)+ as.factor(edu4)+ pct_black + MEDHH +pct_poverty+pct_NoHS+ as.factor(smoke2)+offset(log(non_na_count)),id = df_trimmed$ExtractID,bayes = TRUE, expnms = mixture, data=df_trimmed ,q=10, family=quasipoisson(link = "log"))

summary(mod_ipw)
```

# Frailty Figure 1
```{r}
# create plot Figure 1
# --- Extract weights ---
pos_df <- as.data.frame(mod_ipw$pos.weights)
neg_df <- as.data.frame(mod_ipw$neg.weights)

pos_df <- tibble(variable = rownames(pos_df), weight = pos_df[[1]], direction = "Positive")
neg_df <- tibble(variable = rownames(neg_df), weight = -neg_df[[1]], direction = "Negative")  # flip sign

# Combine and clean
weights_df <- bind_rows(pos_df, neg_df) %>%
    filter(weight != 0)

# --- Clean labels (with math formatting) ---
weights_df$variable_label <- recode(weights_df$variable,
                                    "pm" = "PM[2.5]~'(μg/m'^3*')'",
                                    "mean_temp" = "'Mean Temperature (°C)'",
                                    "Mean_GVI_300" = "'Green View Index (300 m)'"
)

# --- Order so PM2.5 appears at TOP after coord_flip() ---
weights_df$variable_label <- factor(
    weights_df$variable_label,
    levels = rev(c(  # reverse order for top-down display
        "PM[2.5]~'(μg/m'^3*')'",
        "'Mean Temperature (°C)'",
        "'Green View Index (300 m)'"
    ))
)

# --- Plot ---
p <- ggplot(weights_df, aes(x = variable_label, y = weight, fill = direction)) +
    geom_bar(stat = "identity", color = "black", width = 0.75, alpha = 0.9) +
    coord_flip() +
    geom_hline(yintercept = 0, color = "gray40", linewidth = 0.6) +
    scale_fill_manual(values = c("Positive" = "#4C566A", "Negative" = "#A7ADBA")) +
    theme_minimal(base_size = 14) +
    labs(
        title = "Frailty Index",
        x = NULL,
        y = "Weight",
        #fill = "Direction"
    ) +
    theme(
        plot.title = element_text(face = "bold", size = 15, hjust = 0.5),
        panel.grid.major.y = element_blank(),
        legend.position = "none"
    ) +
    scale_x_discrete(labels = function(x) parse(text = x))
```

# Mixture analysis: PTSD # Table 3
```{r}
# --- Run outcome model ---
mix_ptsd = qgcomp.noboot(
    as.numeric(PTSD) ~  
        pm + Mean_GVI_300 + mean_temp + age + as.factor(year) +
        as.factor(Race4) + as.factor(edu4) + pct_black + MEDHH +
        pct_poverty + pct_NoHS + as.factor(work2) + as.factor(smoke2)+ as.factor(Gender),
    id = df_trimmed2$ExtractID,
    bayes = TRUE,
    expnms = mixture,
    data = df_trimmed2,
    q = 10,
    family = quasipoisson(link = "log"))

summary(mix_ptsd)
```

# PTSD Figure 1
```{r}
# Figure 1
# --- Extract weights ---
pos_df <- as.data.frame(mix_ptsd$pos.weights)
neg_df <- as.data.frame(mix_ptsd$neg.weights)

pos_df <- tibble(variable = rownames(pos_df), weight = pos_df[[1]], direction = "Positive")
neg_df <- tibble(variable = rownames(neg_df), weight = -neg_df[[1]], direction = "Negative")  # flip sign

# Combine and clean
weights_df <- bind_rows(pos_df, neg_df) %>%
    filter(weight != 0)

# --- Clean labels (with math formatting) ---
weights_df$variable_label <- recode(weights_df$variable,
                                    "pm" = "PM[2.5]~'(μg/m'^3*')'",
                                    "mean_temp" = "'Mean Temperature (°C)'",
                                    "Mean_GVI_300" = "'Green View Index (300 m)'"
)

# --- Order so PM2.5 appears at TOP after coord_flip() ---
weights_df$variable_label <- factor(
    weights_df$variable_label,
    levels = rev(c(  # reverse order for top-down display
        "PM[2.5]~'(μg/m'^3*')'",
        "'Mean Temperature (°C)'",
        "'Green View Index (300 m)'"
    ))
)

# --- Plot ---
p1 <- ggplot(weights_df, aes(x = variable_label, y = weight, fill = direction)) +
    geom_bar(stat = "identity", color = "black", width = 0.75, alpha = 0.9) +
    coord_flip() +
    geom_hline(yintercept = 0, color = "gray40", linewidth = 0.6) +
    scale_fill_manual(values = c("Positive" = "#4C566A", "Negative" = "#A7ADBA")) +
    theme_minimal(base_size = 14) +
    labs(
        title = "PTSD",
        x = NULL,
        y = "Weight",
        fill = "Direction"
    ) +
    theme(
        plot.title = element_text(face = "bold", size = 15, hjust = 0.5),
        panel.grid.major.y = element_blank(),
        legend.position = "bottom"
    ) +
    scale_x_discrete(labels = function(x) parse(text = x))

```


# Mixture analysis: Depression # Table 3

```{r}
#Outcome model
mix_dep = qgcomp.noboot(as.numeric(depression) ~  
                      pm +  Mean_GVI_300 +mean_temp +age+as.factor(year) +  as.factor(Race4) + as.factor(edu4)+ pct_black + MEDHH + as.factor(Gender)+ pct_poverty+pct_NoHS+as.factor(work2)+as.factor(smoke2),id = df_trimmed2$ExtractID,
                      bayes = TRUE, expnms = mixture, data=df_trimmed2, q=10, family=quasipoisson(link = "log"))

summary(mix_dep)
```

# Depression Figure 1
```{r}

# --- Extract weights ---
pos_df <- as.data.frame(mix_dep$pos.weights)
neg_df <- as.data.frame(mix_dep$neg.weights)

pos_df <- tibble(variable = rownames(pos_df), weight = pos_df[[1]], direction = "Positive")
neg_df <- tibble(variable = rownames(neg_df), weight = -neg_df[[1]], direction = "Negative")  # flip sign

# Combine and clean
weights_df <- bind_rows(pos_df, neg_df) %>%
    filter(weight != 0)

# --- Clean labels (with math formatting) ---
weights_df$variable_label <- recode(weights_df$variable,
                                    "pm" = "PM[2.5]~'(μg/m'^3*')'",
                                    "mean_temp" = "'Mean Temperature (°C)'",
                                    "Mean_GVI_300" = "'Green View Index (300 m)'"
)

# --- Order so PM2.5 appears at TOP after coord_flip() ---
weights_df$variable_label <- factor(
    weights_df$variable_label,
    levels = rev(c(  # reverse order for top-down display
        "PM[2.5]~'(μg/m'^3*')'",
        "'Mean Temperature (°C)'",
        "'Green View Index (300 m)'"
    ))
)

# --- Plot ---
p2 <- ggplot(weights_df, aes(x = variable_label, y = weight, fill = direction)) +
    geom_bar(stat = "identity", color = "black", width = 0.75, alpha = 0.9) +
    coord_flip() +
    geom_hline(yintercept = 0, color = "gray40", linewidth = 0.6) +
    scale_fill_manual(values = c("Positive" = "#4C566A", "Negative" = "#A7ADBA")) +
    theme_minimal(base_size = 14) +
    labs(
        title = "Depression",
        x = NULL,
        y = "Weight",
        fill = "Direction"
    ) +
    theme(
        plot.title = element_text(face = "bold", size = 15, hjust = 0.5),
        panel.grid.major.y = element_blank(),
        legend.position = "none"
    ) +
    scale_x_discrete(labels = function(x) parse(text = x))
```

# merge plots -Figure 1
```{r}
library(gridExtra)

grid.arrange(p,p1,p2,ncol=2)

```
