---
title: "Untitled"
output: html_document
date: "2026-07-24"
---

```{r include=FALSE}
# install required pacakges
install.packages(plyr)
install.packages(dplyr)
install.packages(mgcv)
install.packages(table1)
install.packages(tableone)
install.packages(qgcomp)
install.packages(qgcompint)
install.packages(lubridate)
install.packages(ggplot2)
install.packages(tidyr)
install.packages(tidyverse)
install.packages(sf)
install.packages(patchwork)
```


# run libraries
```{r include=FALSE}
library(plyr)
library(dplyr)
library(mgcv)
library(table1)
library(tableone)
library(qgcomp)
library(qgcompint)
library(lubridate)
library(ggplot2)
library(tidyr)
library(tidyverse)
library(sf)
library(patchwork)
```


save the data to a specific location

# Set working direcory
```{r}
wd = "C:/......"

```

# Import data

List of variables:
# year - monitoring year
# Gender- Male/Female
# Race4- Race in 4 categories
# edu4 -  Education level in 4 categories
# work2 - Occupation on 9/11
# exp911_imp2 - WTC exposure index (High/Very High=1/ Low/Intermediate=0)
# arr_9_11_13 - Arrival date (Arrived on 9/11-9/14 =1/ Arrived after 9/14 =0)
# horror - Witnessing horror (yes=1/ No=0)
# smoke2 -  Smoking status
# age - Age at monitoring visit
# non_na_count - Number of counted deficits
# pct_black - % Black population 
# pct_poverty - % under the poverty line
# pct_NoHS - % with no high school education
# MEDHH - Median household income 
# pm - PM2.5 (µg m-3)
# mean_temp - Mean temperature (° C)
# Mean_GVI_300 - Green View Index inside 300m buffer
# Mean_GVI_500 - Green View Index inside 500m buffer
# frailty.sum - Frailty index is calculated using 30 items that are collected during each monitoring visit, including deficits such as diabetes, fair/poor self-rated health, difficulty climbing several flights of stairs, abnormal heart examination, and self-reported mental health conditions. Each item is coded as a binary value where a response of Fair/Poor is considered as a “deficit” (and assigned a score of 1), and a response of Excellent/Very Good/Good is assigned a score of 0 (i.e., not a deficit). 
# PTSD - PTSD score based on the DSM-IV measurement 
# depression - Depression score based on PHQ9.

```{r}
data = read.csv(paste0(wd, '/Sim_Data.csv'))
head(data)
```

# Subset data for frailty analysis. The population for the frailty analysis is restricted to rows with at least 27 deficits. 
#Rows with less than 27 deficite are considered missing due to insufficient data to calculate the frailty index.  
```{r}
data_index= subset(data, non_na_count >=27)
```

# Remove outliers from the datasets 
# df_trimmed = The data for the frailty analysis
# df_trimmed2 = The data for the mental illness analyses
```{r}
#Function to remove lower and upper 0.5% in the exposure distributions to avoid bias due to extreme values. 
trim_df <- function(data, vars, p = 0.005) {
  # Loop over variables and compute quantiles
  for (v in vars) {
    lower <- quantile(data[[v]], p, na.rm = TRUE)
    upper <- quantile(data[[v]], 1 - p, na.rm = TRUE)
    
    # Keep only rows within range
    data <- data[data[[v]] >= lower & data[[v]] <= upper, ]
  }
  return(data)
}

# Apply to temperature and PM exposures
df_trimmed <- trim_df(data_index, vars = c('pm','mean_temp'))

df_trimmed <- df_trimmed %>%
  filter(!if_all(everything(), is.na))

df_trimmed2 <- trim_df(data, vars = c('pm','mean_temp'))

df_trimmed2 <- df_trimmed2 %>%
  filter(!if_all(everything(), is.na))
```

# Table S5. Sensitivity analysis restricting addresses only from 2012
#Create subsets for the farilty and PTSD/depression analyses
```{r}
subset_frail = subset(df_trimmed,year>=2012)

subset_ptsd = subset(df_trimmed2,year>=2012)
```

#Model for frailty
```{r}
frail_sub <- bam(frailty.sum~ pm +mean_temp+ Mean_GVI_300 +age+s(year,bs='ps') + as.factor(Gender) + as.factor(Race4) + as.factor(edu4)+ pct_black + MEDHH + pct_poverty+pct_NoHS+as.factor(work2)+ as.factor(smoke2)+s(RunningID,bs='re'),data=subset_frail, family = quasipoisson(link = log),offset = log(non_na_count))

    pm=summary(frail_sub)$p.table
    pm<-as.data.frame(pm,keep.rownames = TRUE)[] # Format as datatable
    table_pm=pm[2:4,]
    
    # get upper and lower bound
    table_pm$lower = table_pm$Estimate - qnorm(0.975) * table_pm$`Std. Error`
    table_pm$upper = table_pm$Estimate + qnorm(0.975) * table_pm$`Std. Error`
    table_pm = table_pm[,c(1,5:6) ]

    # PM row
    table_pm$Estimate_exp <- (exp(table_pm$Estimate*iqrs)-1)*100
    table_pm$lower_exp <- (exp(table_pm$lower*iqrs)-1)*100
    table_pm$upper_exp <- (exp(table_pm$upper*iqrs)-1)*100
    
  # Round values 
  table_pm$Estimate_exp <- round(table_pm$Estimate_exp, 3)
  table_pm$lower_exp <- round(table_pm$lower_exp, 3)
  table_pm$upper_exp <- round(table_pm$upper_exp, 3)
  table_pm$complete <- paste(table_pm$Estimate_exp, "(", table_pm$lower_exp, ";", table_pm$upper_exp, ")")
  table_pm[, 7, drop = FALSE]
```

#Model for PTSD
```{r}
#Outcome model
ptsd_sub <- bam(as.numeric(PTSD)~ pm +mean_temp+  Mean_GVI_300  +age+s(year,bs='ps') + as.factor(Gender) + as.factor(Race4) + as.factor(edu4)+ pct_black + MEDHH + pct_poverty+pct_NoHS+as.factor(work2)+ as.factor(smoke2)+s(RunningID,bs='re'),family = quasipoisson(link = log), data=subset_ptsd)

    pm=summary(ptsd_sub)$p.table
    pm<-as.data.frame(pm,keep.rownames = TRUE)[] # Format as datatable
    table_pm=pm[2:4,]
    
    # get upper and lower bound
    table_pm$lower = table_pm$Estimate - qnorm(0.975) * table_pm$`Std. Error`
    table_pm$upper = table_pm$Estimate + qnorm(0.975) * table_pm$`Std. Error`
    table_pm = table_pm[,c(1,5:6) ]

    # PM row
    table_pm$Estimate_exp <- (exp(table_pm$Estimate*iqrs)-1)*100
    table_pm$lower_exp <- (exp(table_pm$lower*iqrs)-1)*100
    table_pm$upper_exp <- (exp(table_pm$upper*iqrs)-1)*100
    
  # Round values 
  table_pm$Estimate_exp <- round(table_pm$Estimate_exp, 3)
  table_pm$lower_exp <- round(table_pm$lower_exp, 3)
  table_pm$upper_exp <- round(table_pm$upper_exp, 3)
  table_pm$complete <- paste(table_pm$Estimate_exp, "(", table_pm$lower_exp, ";", table_pm$upper_exp, ")")
  table_pm[, 7, drop = FALSE]

```
#Model for depression
```{r}
PHQ9_sub <- bam(as.numeric(depression)~ pm +mean_temp+  Mean_GVI_300  +age+s(year,bs='ps') + as.factor(Gender) + as.factor(Race4) + as.factor(edu4)+ pct_black + MEDHH + pct_poverty+pct_NoHS+as.factor(work2)+ as.factor(smoke2)+s(RunningID,bs='re'),family = quasipoisson(link = log),data=subset_ptsd) 

    pm=summary(PHQ9_sub)$p.table
    pm<-as.data.frame(pm,keep.rownames = TRUE)[] # Format as datatable
    table_pm=pm[2:4,]
    
    table_pm$lower = table_pm$Estimate - qnorm(0.975) * table_pm$`Std. Error`
    table_pm$upper = table_pm$Estimate + qnorm(0.975) * table_pm$`Std. Error`
    table_pm = table_pm[,c(1,5:6) ]

    # PM row
    table_pm$Estimate_exp <- (exp(table_pm$Estimate*iqrs)-1)*100
    table_pm$lower_exp <- (exp(table_pm$lower*iqrs)-1)*100
    table_pm$upper_exp <- (exp(table_pm$upper*iqrs)-1)*100
    
  # Round values 
  table_pm$Estimate_exp <- round(table_pm$Estimate_exp, 3)
  table_pm$lower_exp <- round(table_pm$lower_exp, 3)
  table_pm$upper_exp <- round(table_pm$upper_exp, 3)
  table_pm$complete <- paste(table_pm$Estimate_exp, "(", table_pm$lower_exp, ";", table_pm$upper_exp, ")")
  table_pm[, 7, drop = FALSE]
```

#Mixture models
```{r}
fi_500 = qgcomp.noboot(frailty.sum ~  
                      pm +  Mean_GVI_300 +mean_temp +age+as.factor(year) + as.factor(work2)+ as.factor(Race4) + as.factor(edu4)+ pct_black + MEDHH +pct_poverty+pct_NoHS+as.factor(Gender)+ as.factor(smoke2)+offset(log(non_na_count)),id = subset_frail$ExtractID,bayes = TRUE, expnms = mixture, data=subset_frail ,q=10, family=quasipoisson(link = "log"))

summary(fi_500)

ptsd_500 = qgcomp.noboot(
    as.numeric(PTSD) ~  
        pm + Mean_GVI_300 + mean_temp + age + as.factor(year) +
        as.factor(Race4) + as.factor(edu4) + pct_black + MEDHH +as.factor(Gender)+
        pct_poverty + pct_NoHS + as.factor(work2) + as.factor(smoke2),
    id = subset_ptsd$ExtractID,
    bayes = TRUE,
    expnms = mixture,
    data = subset_ptsd,
    q = 10,
    family=quasipoisson(link = "log")
)
summary(ptsd_500)

dep_500 = qgcomp.noboot(as.numeric(depression) ~  
                      pm +  Mean_GVI_300 +mean_temp +age+as.factor(year) +  as.factor(Race4) + as.factor(edu4)+ pct_black + MEDHH +  pct_poverty+pct_NoHS+as.factor(work2)+as.factor(Gender)+as.factor(smoke2),id = subset_ptsd$ExtractID,
                      bayes = TRUE, expnms = mixture, data=subset_ptsd, q=10, family=quasipoisson(link = "log"))

summary(dep_500)

```
