---
title: "Untitled"
output: html_document
date: "2026-07-24"
---

```{r include=FALSE}
# install required pacakges
install.packages(plyr)
install.packages(dplyr)
install.packages(mgcv)
install.packages(table1)
install.packages(tableone)
install.packages(qgcomp)
install.packages(qgcompint)
install.packages(lubridate)
install.packages(ggplot2)
install.packages(tidyr)
install.packages(tidyverse)
install.packages(sf)
install.packages(patchwork)
```


# run libraries
```{r include=FALSE}
library(plyr)
library(dplyr)
library(mgcv)
library(table1)
library(tableone)
library(qgcomp)
library(qgcompint)
library(lubridate)
library(ggplot2)
library(tidyr)
library(tidyverse)
library(sf)
library(patchwork)
```


save the data to a specific location

# Set working direcory
```{r}
wd = "C:/......"

```

# Import data

List of variables:
# year - monitoring year
# Gender- Male/Female
# Race4- Race in 4 categories
# edu4 -  Education level in 4 categories
# work2 - Occupation on 9/11
# exp911_imp2 - WTC exposure index (High/Very High=1/ Low/Intermediate=0)
# arr_9_11_13 - Arrival date (Arrived on 9/11-9/14 =1/ Arrived after 9/14 =0)
# horror - Witnessing horror (yes=1/ No=0)
# smoke2 -  Smoking status
# age - Age at monitoring visit
# non_na_count - Number of counted deficits
# pct_black - % Black population 
# pct_poverty - % under the poverty line
# pct_NoHS - % with no high school education
# MEDHH - Median household income 
# pm - PM2.5 (µg m-3)
# mean_temp - Mean temperature (° C)
# Mean_GVI_300 - Green View Index inside 300m buffer
# Mean_GVI_500 - Green View Index inside 500m buffer
# frailty.sum - Frailty index is calculated using 30 items that are collected during each monitoring visit, including deficits such as diabetes, fair/poor self-rated health, difficulty climbing several flights of stairs, abnormal heart examination, and self-reported mental health conditions. Each item is coded as a binary value where a response of Fair/Poor is considered as a “deficit” (and assigned a score of 1), and a response of Excellent/Very Good/Good is assigned a score of 0 (i.e., not a deficit). 
# PTSD - PTSD score based on the DSM-IV measurement 
# depression - Depression score based on PHQ9.

```{r}
data = read.csv(paste0(wd, '/Sim_Data.csv'))
head(data)
```

#Table S4. The association between the exposures and Frailty: Sensitivity analysis incorporating the probability of having at least 27 items for the index calculation 
# Since 21% of responders had missing FI values, we added IPW and repeated the model as a sensitivity analysis. 
```{r}
#(1) Run logistic model to predict the probability of inclusion in our frailty analysis

# mark whether each row meets the “≥27 non-missing values”
data <- data %>%
  mutate(has27 = as.integer(non_na_count >= 27))

# model the flagged variable
prob_mod <- glm(
  has27 ~ age + as.factor(Gender) + as.factor(Race4) +
    as.factor(edu4) + pct_black + MEDHH + pct_poverty + pct_NoHS +
    as.factor(work2) + as.factor(smoke2),
  data = data,
  family = binomial(link = "logit")
)

# Predicted probability of having ≥27 non-NA values
prob_has27 <- predict(prob_mod, data, type = "response")
data$ipw_27 <- 1 / prob_has27

# take only rows that had more than 27 variables
data_ipw= subset(data, non_na_count >=27)

# remove outliers
data_ipw2 <- trim_df(data_ipw, vars = c('pm','mean_temp'))
data_ipw2 <- data_ipw2 %>%
  filter(!if_all(everything(), is.na))

```

```{r}
# Run the outcome model incorporating the IPW weights
pm_mean <-bam(frailty.sum~ pm +mean_temp+ Mean_GVI_300 +age+s(year,bs='ps') + as.factor(Gender) + as.factor(Race4) + as.factor(edu4)+ pct_black + MEDHH + pct_poverty+pct_NoHS+as.factor(work2)+ as.factor(smoke2)+s(RunningID,bs='re'),data=data_ipw2, family = quasipoisson(link = log),offset = log(non_na_count),weights = ipw_27)

    pm=summary(pm_mean)$p.table
    pm<-as.data.frame(pm,keep.rownames = TRUE)[] # Format as datatable
    table_pm=pm[2:4,]
    
    # get upper and lower bound
    table_pm$lower = table_pm$Estimate - qnorm(0.975) * table_pm$`Std. Error`
    table_pm$upper = table_pm$Estimate + qnorm(0.975) * table_pm$`Std. Error`
    table_pm = table_pm[,c(1,5:6) ]

    # PM row
    table_pm$Estimate_exp <- (exp(table_pm$Estimate*iqrs)-1)*100
    table_pm$lower_exp <- (exp(table_pm$lower*iqrs)-1)*100
    table_pm$upper_exp <- (exp(table_pm$upper*iqrs)-1)*100
    
  # Round values 
  table_pm$Estimate_exp <- round(table_pm$Estimate_exp, 3)
  table_pm$lower_exp <- round(table_pm$lower_exp, 3)
  table_pm$upper_exp <- round(table_pm$upper_exp, 3)
  table_pm$complete <- paste(table_pm$Estimate_exp, "(", table_pm$lower_exp, ";", table_pm$upper_exp, ")")
  table_pm[, 7, drop = FALSE]
```

# mixture with IPW
```{r}
# Outcome model for frailty (qgcomp with IPW)
mod_ipw = qgcomp.noboot(frailty.sum ~  
                      pm +  Mean_GVI_300 +mean_temp +age+as.factor(year) + as.factor(work2)+ as.factor(Race4) + as.factor(edu4)+ pct_black + MEDHH +pct_poverty+pct_NoHS+as.factor(Gender)+ as.factor(smoke2)+offset(log(non_na_count)),id = data_ipw2$RunningID,bayes = TRUE, expnms = mixture, data=data_ipw2 ,q=10, family=quasipoisson(link = "log"),weights = ipw_27)

summary(mod_ipw)
```
 