---
title: "Table 4: effect modification"
output: html_document
date: "2026-04-05"
---
This code represents the original analytic workflow used in the study. Since the original data cannot be shared due to data-use and confidentiality restrictions, we created a simulated dataset to demonstrate code reproducibility and analytic procedures. As a result, the outputs generated from the simulated data differ from the results presented in the manuscript.

The simulation data is located in the zip file attached to this code

The analysis for Table 4 is computationally intensive. Therefore, to facilitate code examination, we added an option to subset 2,000 observations and run the code over 3 bootstrap iterations. In the paper, we used the complete original dataset with b = 1,000 bootstrap iterations.

```{r include=FALSE}
install.packages(dplyr)
install.packages(mgcv)
install.packages(table1)
install.packages(tableone)
install.packages(qgcomp)
install.packages(qgcompint)
install.packages(lubridate)
install.packages(ggplot2)
install.packages(tidyr)
```

#Libraries
```{r include=FALSE}
library(dplyr)
library(mgcv)
library(table1)
library(tableone)
library(qgcomp)
library(qgcompint)
library(lubridate)
library(ggplot2)
library(tidyr)
```

# set working direcory
```{r}
#wd = "C:/Users/furmah01/OneDrive - The Mount Sinai Hospital/WTC new data 2022/Frailty/Manuscript/After review/data"
wd = "C:/Users/yitshm01/OneDrive - The Mount Sinai Hospital/Maayans lab/Pablo/WTC new data 2022/Frailty/Manuscript/After review/data"
```

# import data
```{r}
data = read.csv(paste0(wd, '/Sim_Data.csv'))
# CODE CHECK ONLY: restrict to 300 observations to shorten processing time; remove for full analysis
set.seed(123)
data <- data[sample(nrow(data), 2000), ]
```

#Binary modifiers as factors
```{r}
# low exposure or intermidiate/high
data$exp911_imp2<- as.factor(data$exp911_imp2)
data$arr_9_11_13<- as.factor(data$arr_9_11_13)
data$horror<- as.factor(data$horror)
```

# The population for the frailty analysis is restricted to rows with less than 27 deficits 
```{r}
data_index= subset(data, non_na_count >=27)
```

#Datasets
# Remove outliers from the full and restricted datasets 
# df_trimmed = The data for the frailty analysis
# df_trimmed2 = The data for the mental illness analyses
```{r}
#Function to remove lower and upper 0.5% in the exposure distributions to avoid bias due to extreme values. 
trim_df <- function(data, vars, p = 0.005) {
  # Loop over variables and compute quantiles
  for (v in vars) {
    lower <- quantile(data[[v]], p, na.rm = TRUE)
    upper <- quantile(data[[v]], 1 - p, na.rm = TRUE)
    
    # Keep only rows within range
    data <- data[data[[v]] >= lower & data[[v]] <= upper, ]
  }
  return(data)
}

# Apply to temperature and PM exposures
df_trimmed <- trim_df(data_index, vars = c('pm','mean_temp'))

df_trimmed <- df_trimmed %>%
  filter(!if_all(everything(), is.na))

df_trimmed2 <- trim_df(data, vars = c('pm','mean_temp'))

df_trimmed2 <- df_trimmed2 %>%
  filter(!if_all(everything(), is.na))
```

#Mixture
```{r}
mixture =  c("pm", "mean_temp", 'Mean_GVI_300')
```

#Complete cases subset for the frailty analysis
```{r}
df_complete <- df_trimmed  %>%
  dplyr::select(RunningID,frailty.sum,non_na_count, pm, mean_temp, Mean_GVI_300, age, year, Race4, edu4,pct_black, MEDHH, pct_poverty, pct_NoHS, work2, smoke2,arr_9_11_13,exp911_imp2,horror,Gender) %>%
  na.omit()
```

#For PTSD
```{r}
df_complete2 <- df_trimmed2  %>%
  dplyr::select(RunningID,PTSD,non_na_count, pm, mean_temp, Mean_GVI_300, age, year, Race4, edu4,pct_black, MEDHH, pct_poverty, pct_NoHS, work2, smoke2,arr_9_11_13,exp911_imp2,horror,Gender) %>%
  na.omit()
```

#For depression
```{r}
df_complete3 <- df_trimmed2 %>%
  dplyr::select(RunningID,depression, pm, mean_temp, Mean_GVI_300, age, year, Race4, edu4,pct_black, MEDHH, pct_poverty, pct_NoHS, work2, exp911_imp2, smoke2, arr_9_11_13, horror,Gender) %>%
  na.omit()
```


#Models for frailty
#WTC index
```{r setup, include=FALSE}
# --- Model 1: exp911_imp2 ---
qfit_exp <- qgcomp.emm.glm.boot(
  frailty.sum ~ pm + Mean_GVI_300 + mean_temp + age +
    as.factor(Gender) + as.factor(year) + as.factor(Race4) + as.factor(edu4) +
    pct_black + MEDHH + pct_poverty + pct_NoHS +
    as.factor(smoke2) + as.factor(work2),
  data = df_complete, expnms = mixture, emmvar = "exp911_imp2",
  q = 10, family = poisson(link = "log"), id = "RunningID",
  B = 3, degree = 1, parallel = TRUE, parplan = TRUE #Use b=1000, changed to 3 to examine the code
)

qfit_exp

# Save just the coefficient table
# Extract coefficient table with CIs
coef_exp <- data.frame(
  term       = names(qfit_exp$coef),
  Estimate   = qfit_exp$coef,
  SE         = sqrt(diag(qfit_exp$covmat.coef)),
  LowerCI    = qfit_exp$ci.coef[, 1],
  UpperCI    = qfit_exp$ci.coef[, 2]
)
coef_exp$Zvalue <- coef_exp$Estimate / coef_exp$SE
coef_exp$Pvalue <- 2 * (1 - pnorm(abs(coef_exp$Zvalue)))

write.csv(coef_exp, paste0(wd, "/frailty_exp911.csv"))
```

# --- Model 2: arr_9_11_13 ---
```{r}
qfit_arr <- qgcomp.emm.glm.boot(
  frailty.sum ~ pm + Mean_GVI_300 + mean_temp + age +
    as.factor(Gender) + as.factor(year) + as.factor(Race4) + as.factor(edu4) +
    pct_black + MEDHH + pct_poverty + pct_NoHS +
    as.factor(smoke2) + as.factor(work2),
  data = df_complete, expnms = mixture, emmvar = "arr_9_11_13",
  q = 10, family = poisson(link = "log"), id ="RunningID",
  B = 3, degree = 1, parallel = TRUE, parplan = TRUE #Use b=1000, changed to 3 to examine the code
)

qfit_arr

# Save just the coefficient table
# Extract coefficient table with CIs
coef_exp <- data.frame(
  term       = names(qfit_arr$coef),
  Estimate   = qfit_arr$coef,
  SE         = sqrt(diag(qfit_arr$covmat.coef)),
  LowerCI    = qfit_arr$ci.coef[, 1],
  UpperCI    = qfit_arr$ci.coef[, 2]
)
coef_exp$Zvalue <- coef_exp$Estimate / coef_exp$SE
coef_exp$Pvalue <- 2 * (1 - pnorm(abs(coef_exp$Zvalue)))

write.csv(coef_exp, paste0(wd, "/frailty_arrival.csv"))
```

```{r}
qfit_hor <- qgcomp.emm.glm.boot(
  frailty.sum ~ pm + Mean_GVI_300 + mean_temp + age +
    as.factor(Gender) + as.factor(year) + as.factor(Race4) + as.factor(edu4) +
    pct_black + MEDHH + pct_poverty + pct_NoHS +
    as.factor(smoke2) + as.factor(work2),
  data = df_complete, expnms = mixture, emmvar = "horror",
  q = 10, family = poisson(link = "log"), id = "RunningID",
  B = 3, degree = 1, parallel = TRUE, parplan = TRUE #Use b=1000, changed to 3 to examine the code
)

qfit_hor

coef_exp <- data.frame(
  term       = names(qfit_hor$coef),
  Estimate   = qfit_hor$coef,
  SE         = sqrt(diag(qfit_hor$covmat.coef)),
  LowerCI    = qfit_hor$ci.coef[, 1],
  UpperCI    = qfit_hor$ci.coef[, 2]
)
coef_exp$Zvalue <- coef_exp$Estimate / coef_exp$SE
coef_exp$Pvalue <- 2 * (1 - pnorm(abs(coef_exp$Zvalue)))

write.csv(coef_exp, paste0(wd, "/frailty_horror.csv"))
```

#Save vcov for the SE calculations
```{r}
# Save covariance matrix alongside coefficients
saveRDS(qfit_exp$covmat.coef, paste0(wd, "/qfit_frailty_exp911_covmat.rds"))
saveRDS(qfit_arr$covmat.coef, paste0(wd, "/qfit_frailty_arr_covmat.rds"))
saveRDS(qfit_hor$covmat.coef, paste0(wd, "/qfit_frailty_hor_covmat.rds"))
```

#Load results
```{r}
make_row_csv <- function(modifier, coef_df, covmat) {
  
  int_term <- coef_df$term[grep("mixture", coef_df$term)]
  
  psi0     <- coef_df$Estimate[coef_df$term == "psi1"]
  ci_low0  <- coef_df$LowerCI[coef_df$term == "psi1"]
  ci_high0 <- coef_df$UpperCI[coef_df$term == "psi1"]
  
  int_est  <- coef_df$Estimate[coef_df$term == int_term]
  p_int    <- coef_df$Pvalue[coef_df$term == int_term]
  
  psi1 <- psi0 + int_est
  
  # Proper SE using full covariance matrix
  se1 <- sqrt(
    covmat["psi1", "psi1"] + 
    covmat[int_term, int_term] + 
    2 * covmat["psi1", int_term]
  )
  
  data.frame(
    modifier      = modifier,
    pct_change_0  = round((exp(psi0) - 1) * 100, 3),
    ci_low_0      = round((exp(ci_low0) - 1) * 100, 3),
    ci_high_0     = round((exp(ci_high0) - 1) * 100, 3),
    pct_change_1  = round((exp(psi1) - 1) * 100, 3),
    ci_low_1      = round((exp(psi1 - 1.96 * se1) - 1) * 100, 3),
    ci_high_1     = round((exp(psi1 + 1.96 * se1) - 1) * 100, 3),
    p_interaction = round(p_int, 4)
  )
}

# Load results
wtc_coef <- read.csv(paste0(wd,"/frailty_exp911.csv"))
arr_coef <- read.csv(paste0(wd,"/frailty_arrival.csv"))
hor_coef <- read.csv(paste0(wd,"/frailty_horror.csv"))

# Load covariance matrices
covmat_exp <- readRDS(paste0(wd,"/qfit_frailty_exp911_covmat.rds"))
covmat_arr <- readRDS(paste0(wd,"/qfit_frailty_arr_covmat.rds"))
covmat_hor <- readRDS(paste0(wd,"/qfit_frailty_hor_covmat.rds"))

results <- bind_rows(
  make_row_csv("exp911_imp2", wtc_coef, covmat_exp),
  make_row_csv("arr_9_11_13", arr_coef, covmat_arr),
  make_row_csv("horror",      hor_coef, covmat_hor)
)

print(results)
```

##########
#Models for ptsd
#WTC index
```{r setup, include=FALSE}
# --- Model 1: exp911_imp2 ---

qfit_exp_ptsd <- qgcomp.emm.glm.boot(
  as.integer(PTSD) ~ pm + Mean_GVI_300 + mean_temp + age +
    as.factor(Gender) + as.factor(year) + as.factor(Race4) + as.factor(edu4) +
    pct_black + MEDHH + pct_poverty + pct_NoHS +
    as.factor(smoke2) + as.factor(work2),
  data = df_complete2, expnms = mixture, emmvar = "exp911_imp2",
  q = 10, family = poisson(link = "log"), id = "RunningID",
  B = 3, degree = 1, parallel = TRUE, parplan = TRUE #Use b=1000, changed to 3 to examine the code
)

qfit_exp_ptsd

coef_exp <- data.frame(
  term       = names(qfit_exp_ptsd$coef),
  Estimate   = qfit_exp_ptsd$coef,
  SE         = sqrt(diag(qfit_exp_ptsd$covmat.coef)),
  LowerCI    = qfit_exp_ptsd$ci.coef[, 1],
  UpperCI    = qfit_exp_ptsd$ci.coef[, 2]
)
coef_exp$Zvalue <- coef_exp$Estimate / coef_exp$SE
coef_exp$Pvalue <- 2 * (1 - pnorm(abs(coef_exp$Zvalue)))

write.csv(coef_exp, paste0(wd,"/ptsd_wtc.csv"))
```

# --- Model 2: arr_9_11_13 ---
```{r}
qfit_arr_ptsd <- qgcomp.emm.glm.boot(
  as.integer(PTSD) ~ pm + Mean_GVI_300 + mean_temp + age +
    as.factor(Gender) + as.factor(year) + as.factor(Race4) + as.factor(edu4) +
    pct_black + MEDHH + pct_poverty + pct_NoHS +
    as.factor(smoke2) + as.factor(work2),
  data = df_complete2, expnms = mixture, emmvar = "arr_9_11_13",
  q = 10, family = poisson(link = "log"), id ="RunningID",
  B = 3, degree = 1, parallel = TRUE, parplan = TRUE #Use b=1000, changed to 3 to examine the code
)

qfit_arr_ptsd

coef_exp <- data.frame(
  term       = names(qfit_arr_ptsd$coef),
  Estimate   = qfit_arr_ptsd$coef,
  SE         = sqrt(diag(qfit_arr_ptsd$covmat.coef)),
  LowerCI    = qfit_arr_ptsd$ci.coef[, 1],
  UpperCI    = qfit_arr_ptsd$ci.coef[, 2]
)
coef_exp$Zvalue <- coef_exp$Estimate / coef_exp$SE
coef_exp$Pvalue <- 2 * (1 - pnorm(abs(coef_exp$Zvalue)))

write.csv(coef_exp, paste0(wd,"/ptsd_arr.csv"))
```

```{r}
qfit_hor_ptsd <- qgcomp.emm.glm.boot(
  as.integer(PTSD) ~ pm + Mean_GVI_300 + mean_temp + age +
    as.factor(Gender) + as.factor(year) + as.factor(Race4) + as.factor(edu4) +
    pct_black + MEDHH + pct_poverty + pct_NoHS +
    as.factor(smoke2) + as.factor(work2),
  data = df_complete2, expnms = mixture, emmvar = "horror",
  q = 10, family = poisson(link = "log"), id = "RunningID",
  B = 3, degree = 1, parallel = TRUE, parplan = TRUE #Use b=1000, changed to 3 to examine the code
)

qfit_hor_ptsd

coef_exp <- data.frame(
  term       = names(qfit_hor_ptsd$coef),
  Estimate   = qfit_hor_ptsd$coef,
  SE         = sqrt(diag(qfit_hor_ptsd$covmat.coef)),
  LowerCI    = qfit_hor_ptsd$ci.coef[, 1],
  UpperCI    = qfit_hor_ptsd$ci.coef[, 2]
)
coef_exp$Zvalue <- coef_exp$Estimate / coef_exp$SE
coef_exp$Pvalue <- 2 * (1 - pnorm(abs(coef_exp$Zvalue)))

write.csv(coef_exp, paste0(wd,"/ptsd_hor.csv"))
```

#Save vcov for the SE calculations
```{r}
# Save covariance matrix alongside coefficients
saveRDS(qfit_exp_ptsd$covmat.coef,paste0(wd, "/ptsd_exp911_covmat.rds"))
saveRDS(qfit_arr_ptsd$covmat.coef, paste0(wd,"/ptsd_arr_covmat.rds"))
saveRDS(qfit_hor_ptsd$covmat.coef,paste0(wd, "/ptsd_hor_covmat.rds"))
```

#Load results
```{r}
make_row_csv <- function(modifier, coef_df, covmat) {
  
  int_term <- coef_df$term[grep("mixture", coef_df$term)]
  
  psi0     <- coef_df$Estimate[coef_df$term == "psi1"]
  ci_low0  <- coef_df$LowerCI[coef_df$term == "psi1"]
  ci_high0 <- coef_df$UpperCI[coef_df$term == "psi1"]
  
  int_est  <- coef_df$Estimate[coef_df$term == int_term]
  p_int    <- coef_df$Pvalue[coef_df$term == int_term]
  
  psi1 <- psi0 + int_est
  
  # Proper SE using full covariance matrix
  se1 <- sqrt(
    covmat["psi1", "psi1"] + 
    covmat[int_term, int_term] + 
    2 * covmat["psi1", int_term]
  )
  
  data.frame(
    modifier      = modifier,
    pct_change_0  = round((exp(psi0) - 1) * 100, 3),
    ci_low_0      = round((exp(ci_low0) - 1) * 100, 3),
    ci_high_0     = round((exp(ci_high0) - 1) * 100, 3),
    pct_change_1  = round((exp(psi1) - 1) * 100, 3),
    ci_low_1      = round((exp(psi1 - 1.96 * se1) - 1) * 100, 3),
    ci_high_1     = round((exp(psi1 + 1.96 * se1) - 1) * 100, 3),
    p_interaction = round(p_int, 4)
  )
}

# Load results
wtc_coef <- read.csv(paste0(wd,"/ptsd_wtc.csv"))
arr_coef <- read.csv(paste0(wd,"/ptsd_arr.csv"))
hor_coef <- read.csv(paste0(wd,"/ptsd_hor.csv"))

# Load covariance matrices
covmat_exp <- readRDS(paste0(wd,"/ptsd_exp911_covmat.rds"))
covmat_arr <- readRDS(paste0(wd,"/ptsd_arr_covmat.rds"))
covmat_hor <- readRDS(paste0(wd,"/ptsd_hor_covmat.rds"))

results <- bind_rows(
  make_row_csv("exp911_imp2", wtc_coef, covmat_exp),
  make_row_csv("arr_9_11_13", arr_coef, covmat_arr),
  make_row_csv("horror",      hor_coef, covmat_hor)
)

print(results)
```



##########
#Models for depression
#WTC index
```{r setup, include=FALSE}
# --- Model 1: exp911_imp2 ---

qfit_exp_dep <- qgcomp.emm.glm.boot(
  as.integer(depression) ~ pm + Mean_GVI_300 + mean_temp + age +
    as.factor(Gender) + as.factor(year) + as.factor(Race4) + as.factor(edu4) +
    pct_black + MEDHH + pct_poverty + pct_NoHS +
    as.factor(smoke2) + as.factor(work2),
  data = df_complete3, expnms = mixture, emmvar = "exp911_imp2",
  q = 10, family = poisson(link = "log"), id = "RunningID",
  B = 3, degree = 1, parallel = TRUE, parplan = TRUE #Use b=1000, changed to 3 to examine the code
)

qfit_exp_dep

coef_exp <- data.frame(
  term       = names(qfit_exp_dep$coef),
  Estimate   = qfit_exp_dep$coef,
  SE         = sqrt(diag(qfit_exp_dep$covmat.coef)),
  LowerCI    = qfit_exp_dep$ci.coef[, 1],
  UpperCI    = qfit_exp_dep$ci.coef[, 2]
)
coef_exp$Zvalue <- coef_exp$Estimate / coef_exp$SE
coef_exp$Pvalue <- 2 * (1 - pnorm(abs(coef_exp$Zvalue)))

write.csv(coef_exp,paste0(wd, "/dep_wtc.csv"))
```

# --- Model 2: arr_9_11_13 ---
```{r}
qfit_arr_dep <- qgcomp.emm.glm.boot(
  as.integer(depression) ~ pm + Mean_GVI_300 + mean_temp + age +
    as.factor(Gender) + as.factor(year) + as.factor(Race4) + as.factor(edu4) +
    pct_black + MEDHH + pct_poverty + pct_NoHS +
    as.factor(smoke2) + as.factor(work2),
  data = df_complete3, expnms = mixture, emmvar = "arr_9_11_13",
  q = 10, family = poisson(link = "log"), id ="RunningID",
  B = 3, degree = 1, parallel = TRUE, parplan = TRUE #Use b=1000, changed to 3 to examine the code
)

qfit_arr_dep

coef_exp <- data.frame(
  term       = names(qfit_arr_dep$coef),
  Estimate   = qfit_arr_dep$coef,
  SE         = sqrt(diag(qfit_arr_dep$covmat.coef)),
  LowerCI    = qfit_arr_dep$ci.coef[, 1],
  UpperCI    = qfit_arr_dep$ci.coef[, 2]
)
coef_exp$Zvalue <- coef_exp$Estimate / coef_exp$SE
coef_exp$Pvalue <- 2 * (1 - pnorm(abs(coef_exp$Zvalue)))

write.csv(coef_exp,paste0(wd, "/dep_arr.csv"))

```

```{r}
qfit_hor_dep <- qgcomp.emm.glm.boot(
  as.integer(depression) ~ pm + Mean_GVI_300 + mean_temp + age +
    as.factor(Gender) + as.factor(year) + as.factor(Race4) + as.factor(edu4) +
    pct_black + MEDHH + pct_poverty + pct_NoHS +
    as.factor(smoke2) + as.factor(work2),
  data = df_complete3, expnms = mixture, emmvar = "horror",
  q = 10, family = poisson(link = "log"), id = "RunningID",
  B = 3, degree = 1, parallel = TRUE, parplan = TRUE #Use b=1000, changed to 3 to examine the code
)

qfit_hor_dep

coef_exp <- data.frame(
  term       = names(qfit_hor_dep$coef),
  Estimate   = qfit_hor_dep$coef,
  SE         = sqrt(diag(qfit_hor_dep$covmat.coef)),
  LowerCI    = qfit_hor_dep$ci.coef[, 1],
  UpperCI    = qfit_hor_dep$ci.coef[, 2]
)
coef_exp$Zvalue <- coef_exp$Estimate / coef_exp$SE
coef_exp$Pvalue <- 2 * (1 - pnorm(abs(coef_exp$Zvalue)))

write.csv(coef_exp,paste0(wd, "/dep_hor.csv"))
```

#Save vcov for the SE calculations
```{r}
# Save covariance matrix alongside coefficients
saveRDS(qfit_exp_dep$covmat.coef, paste0(wd,"/dep_exp911_covmat.rds"))
saveRDS(qfit_arr_dep$covmat.coef, paste0(wd,"/dep_arr_covmat.rds"))
saveRDS(qfit_hor_dep$covmat.coef,paste0(wd, "/dep_hor_covmat.rds"))
```

```{r}
make_row_csv <- function(modifier, coef_df, covmat) {
  
  int_term <- coef_df$term[grep("mixture", coef_df$term)]
  
  psi0     <- coef_df$Estimate[coef_df$term == "psi1"]
  ci_low0  <- coef_df$LowerCI[coef_df$term == "psi1"]
  ci_high0 <- coef_df$UpperCI[coef_df$term == "psi1"]
  
  int_est  <- coef_df$Estimate[coef_df$term == int_term]
  p_int    <- coef_df$Pvalue[coef_df$term == int_term]
  
  psi1 <- psi0 + int_est
  
  # Proper SE using full covariance matrix
  se1 <- sqrt(
    covmat["psi1", "psi1"] + 
    covmat[int_term, int_term] + 
    2 * covmat["psi1", int_term]
  )
  
  data.frame(
    modifier      = modifier,
    pct_change_0  = round((exp(psi0) - 1) * 100, 3),
    ci_low_0      = round((exp(ci_low0) - 1) * 100, 3),
    ci_high_0     = round((exp(ci_high0) - 1) * 100, 3),
    pct_change_1  = round((exp(psi1) - 1) * 100, 3),
    ci_low_1      = round((exp(psi1 - 1.96 * se1) - 1) * 100, 3),
    ci_high_1     = round((exp(psi1 + 1.96 * se1) - 1) * 100, 3),
    p_interaction = round(p_int, 4)
  )
}

# Load results
wtc_coef <- read.csv(paste0(wd,"/dep_wtc.csv"))
arr_coef <- read.csv(paste0(wd,"/dep_arr.csv"))
hor_coef <- read.csv(paste0(wd,"/dep_hor.csv"))

# Load covariance matrices
covmat_exp <- readRDS(paste0(wd,"/dep_exp911_covmat.rds"))
covmat_arr <- readRDS(paste0(wd,"/dep_arr_covmat.rds"))
covmat_hor <- readRDS(paste0(wd,"/dep_hor_covmat.rds"))

results <- bind_rows(
  make_row_csv("exp911_imp2", wtc_coef, covmat_exp),
  make_row_csv("arr_9_11_13", arr_coef, covmat_arr),
  make_row_csv("horror",      hor_coef, covmat_hor)
)

print(results)
```

