---
title: "Untitled"
output: html_document
date: "2026-06-02"
---

```{r include=FALSE}
# install required pacakges
install.packages(tidyverse)
install.packages(dplyr)
install.packages(mgcv)
install.packages(lubridate)
install.packages(ggplot2)
install.packages(tidyr)
install.packages(patchwork)
```

```{r include=FALSE}
library(dplyr)
library(mgcv)
library(lubridate)
library(ggplot2)
library(tidyr)
library(tidyverse)
library(patchwork)
```

# set working direcory
```{r}
wd = "C:/......"
```

```{r}
data = read.csv(paste0(wd, '/Sim_Data.csv'))
```

# Import data

List of variables:
# year - monitoring year
# Gender- Male/Female
# Race4- Race in 4 categories
# edu4 -  Education level in 4 categories
# work2 - Occupation on 9/11
# exp911_imp2 - WTC exposure index (High/Very High=1/ Low/Intermediate=0)
# arr_9_11_13 - Arrival date (Arrived on 9/11-9/14 =1/ Arrived after 9/14 =0)
# horror - Witnessing horror (yes=1/ No=0)
# smoke2 -  Smoking status
# age - Age at monitoring visit
# non_na_count - Number of counted deficits
# pct_black - % Black population 
# pct_poverty - % under the poverty line
# pct_NoHS - % with no high school education
# MEDHH - Median household income 
# pm - PM2.5 (µg m-3)
# mean_temp - Mean temperature (° C)
# Mean_GVI_300 - Green View Index inside 300m buffer
# Mean_GVI_500 - Green View Index inside 500m buffer
# frailty.sum - Frailty index is calculated using 30 items that are collected during each monitoring visit, including deficits such as diabetes, fair/poor self-rated health, difficulty climbing several flights of stairs, abnormal heart examination, and self-reported mental health conditions. Each item is coded as a binary value where a response of Fair/Poor is considered as a “deficit” (and assigned a score of 1), and a response of Excellent/Very Good/Good is assigned a score of 0 (i.e., not a deficit). 
# PTSD - PTSD score based on the DSM-IV measurement 
# depression - Depression score based on PHQ9.

# file containes both winter and summer values in separte columns
```{r}
temp = read.csv(paste0(wd, '/Sim_Summer_temp.csv'))

temp2 = cbind(data, temp)

# clean
temp2$X=NULL
```

# check correlations between the temperatures
```{r}
# Install patchwork if you don't have it: install.packages("patchwork")
library(ggplot2)
library(patchwork)

# --- PLOT 1 ---
# Calculate R2 for your first dataset/subset
r2_1 <- summary(lm(temp_summer ~ mean_temp, data = temp2))$r.squared
label_1 <- paste0("italic(R)^2 == ", round(r2_1, 3))

p1 <- ggplot(temp2, aes(x = mean_temp, y = temp_summer)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  annotate("text", x = -Inf, y = Inf, label = label_1, 
           parse = TRUE, hjust = -0.2, vjust = 1.5, size = 5) +
  theme_minimal()

# --- PLOT 2 ---
# Calculate R2 for your second dataset/subset
r2_2 <- summary(lm(temp_winter ~ mean_temp, data = temp2))$r.squared
label_2 <- paste0("italic(R)^2 == ", round(r2_2, 3))

p2 <- ggplot(temp2, aes(x = mean_temp, y = temp_summer)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  annotate("text", x = -Inf, y = Inf, label = label_2, 
           parse = TRUE, hjust = -0.2, vjust = 1.5, size = 5) +
  theme_minimal()

p3 <- ggplot(temp2, aes(x = temp_winter, y = temp_summer)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  annotate("text", x = -Inf, y = Inf, label = label_2, 
           parse = TRUE, hjust = -0.2, vjust = 1.5, size = 5) +
  theme_minimal()

# --- SIDE BY SIDE ---
# Patchwork lets you just "add" them together for a side-by-side view
p1 + p2+p3
```


# The population for the frailty analysis is restricted to rows with less than 27 deficits 
```{r}
data_index= subset(temp2, non_na_count >=27)
```


######################### summer #############################################

# Remove outliers from the full and restricted datasets 
# df_trimmed = The data for the frailty analysis
# df_trimmed2 = The data for the mental illness analyses
```{r}
#Function to remove lower and upper 0.5% in the exposure distributions to avoid bias due to extreme values. 
trim_df <- function(data, vars, p = 0.005) {
  # Loop over variables and compute quantiles
  for (v in vars) {
    lower <- quantile(data[[v]], p, na.rm = TRUE)
    upper <- quantile(data[[v]], 1 - p, na.rm = TRUE)
    
    # Keep only rows within range
    data <- data[data[[v]] >= lower & data[[v]] <= upper, ]
  }
  return(data)
}

# Apply to temperature and PM exposures
df_trimmed <- trim_df(data_index, vars = c('pm','temp_summer'))

df_trimmed <- df_trimmed %>%
  filter(!if_all(everything(), is.na))

df_trimmed2 <- trim_df(temp2, vars = c('pm','temp_summer'))

df_trimmed2 <- df_trimmed2 %>%
  filter(!if_all(everything(), is.na))
```

# Compute IQR for the main exposures 
```{r}
var <- c("pm", "temp_summer", "Mean_GVI_300")

iqrs <- sapply(var, function(var) IQR(temp2[[var]], na.rm = TRUE))
```

# Model 1: Frialty analysis
```{r}
frail_mod <- bam(frailty.sum~ pm +temp_summer+ Mean_GVI_300 +age+s(year,bs='ps') + as.factor(Gender) + as.factor(Race4) + as.factor(edu4)+ pct_black + MEDHH + pct_poverty+pct_NoHS+as.factor(work2)+ as.factor(smoke2)+s(RunningID,bs='re'),data=df_trimmed, family = quasipoisson(link = log),offset = log(non_na_count))

    pm=summary(frail_mod)$p.table
    pm<-as.data.frame(pm,keep.rownames = TRUE)[] # Format as datatable
    table_pm=pm[2:4,]
    
    # get upper and lower bound
    table_pm$lower = table_pm$Estimate - qnorm(0.975) * table_pm$`Std. Error`
    table_pm$upper = table_pm$Estimate + qnorm(0.975) * table_pm$`Std. Error`
    table_pm = table_pm[,c(1,5:6) ]

    # PM row
    table_pm$Estimate_exp <- (exp(table_pm$Estimate*iqrs)-1)*100
    table_pm$lower_exp <- (exp(table_pm$lower*iqrs)-1)*100
    table_pm$upper_exp <- (exp(table_pm$upper*iqrs)-1)*100
    
  # Round values 
  table_pm$Estimate_exp <- round(table_pm$Estimate_exp, 3)
  table_pm$lower_exp <- round(table_pm$lower_exp, 3)
  table_pm$upper_exp <- round(table_pm$upper_exp, 3)
  table_pm$complete <- paste(table_pm$Estimate_exp, "(", table_pm$lower_exp, ";", table_pm$upper_exp, ")")
  table_pm[, 7, drop = FALSE]
```

# Model 2: PTSD analysis
```{r}
#Outcome model
ptsd_mod <- bam(as.numeric(PTSD)~ pm +temp_summer+ Mean_GVI_300 +age+s(year,bs='ps') + as.factor(Gender) + as.factor(Race4) + as.factor(edu4)+ pct_black + MEDHH + pct_poverty+pct_NoHS+as.factor(work2)+ as.factor(smoke2)+s(RunningID,bs='re'),family = quasipoisson(link = log), data=df_trimmed2)

    pm=summary(ptsd_mod)$p.table
    pm<-as.data.frame(pm,keep.rownames = TRUE)[] # Format as datatable
    table_pm=pm[2:4,]
    
    # get upper and lower bound
    table_pm$lower = table_pm$Estimate - qnorm(0.975) * table_pm$`Std. Error`
    table_pm$upper = table_pm$Estimate + qnorm(0.975) * table_pm$`Std. Error`
    table_pm = table_pm[,c(1,5:6) ]

    # PM row
    table_pm$Estimate_exp <- (exp(table_pm$Estimate*iqrs)-1)*100
    table_pm$lower_exp <- (exp(table_pm$lower*iqrs)-1)*100
    table_pm$upper_exp <- (exp(table_pm$upper*iqrs)-1)*100
    
  # Round values 
  table_pm$Estimate_exp <- round(table_pm$Estimate_exp, 3)
  table_pm$lower_exp <- round(table_pm$lower_exp, 3)
  table_pm$upper_exp <- round(table_pm$upper_exp, 3)
  table_pm$complete <- paste(table_pm$Estimate_exp, "(", table_pm$lower_exp, ";", table_pm$upper_exp, ")")
  table_pm[, 7, drop = FALSE]

```

# Model 3 - Depression outcome 0-9
```{r}
PHQ9_mod <- bam(as.numeric(depression)~ pm +temp_summer+ Mean_GVI_300 +age+s(year,bs='ps') + as.factor(Gender) + as.factor(Race4) + as.factor(edu4)+ pct_black + MEDHH + pct_poverty+pct_NoHS+as.factor(work2)+ as.factor(smoke2)+s(RunningID,bs='re'),family = quasipoisson(link = log),data=df_trimmed2) 

    pm=summary(PHQ9_mod)$p.table
    pm<-as.data.frame(pm,keep.rownames = TRUE)[] # Format as datatable
    table_pm=pm[2:4,]
    
    table_pm$lower = table_pm$Estimate - qnorm(0.975) * table_pm$`Std. Error`
    table_pm$upper = table_pm$Estimate + qnorm(0.975) * table_pm$`Std. Error`
    table_pm = table_pm[,c(1,5:6) ]

    # PM row
    table_pm$Estimate_exp <- (exp(table_pm$Estimate*iqrs)-1)*100
    table_pm$lower_exp <- (exp(table_pm$lower*iqrs)-1)*100
    table_pm$upper_exp <- (exp(table_pm$upper*iqrs)-1)*100
    
  # Round values 
  table_pm$Estimate_exp <- round(table_pm$Estimate_exp, 3)
  table_pm$lower_exp <- round(table_pm$lower_exp, 3)
  table_pm$upper_exp <- round(table_pm$upper_exp, 3)
  table_pm$complete <- paste(table_pm$Estimate_exp, "(", table_pm$lower_exp, ";", table_pm$upper_exp, ")")
  table_pm[, 7, drop = FALSE]

```

######################### winter ########################
# Remove outliers from the full and restricted datasets 
# df_trimmed = The data for the frailty analysis
# df_trimmed2 = The data for the mental illness analyses
```{r}
#Function to remove lower and upper 0.5% in the exposure distributions to avoid bias due to extreme values. 
trim_df <- function(data, vars, p = 0.005) {
  # Loop over variables and compute quantiles
  for (v in vars) {
    lower <- quantile(data[[v]], p, na.rm = TRUE)
    upper <- quantile(data[[v]], 1 - p, na.rm = TRUE)
    
    # Keep only rows within range
    data <- data[data[[v]] >= lower & data[[v]] <= upper, ]
  }
  return(data)
}

# Apply to temperature and PM exposures
df_trimmed <- trim_df(data_index, vars = c('pm','temp_winter'))

df_trimmed <- df_trimmed %>%
  filter(!if_all(everything(), is.na))

df_trimmed2 <- trim_df(temp2, vars = c('pm','temp_winter'))

df_trimmed2 <- df_trimmed2 %>%
  filter(!if_all(everything(), is.na))
```

# Compute IQR for the main exposures 
```{r}
var <- c("pm", "temp_winter", "Mean_GVI_300")

iqrs <- sapply(var, function(var) IQR(temp2[[var]], na.rm = TRUE))
```

# Model 1: Frialty analysis
```{r}
frail_mod <- bam(frailty.sum~ pm +temp_winter+ Mean_GVI_300 +age+s(year,bs='ps') + as.factor(Gender) + as.factor(Race4) + as.factor(edu4)+ pct_black + MEDHH + pct_poverty+pct_NoHS+as.factor(work2)+ as.factor(smoke2)+s(RunningID,bs='re'),data=df_trimmed, family = quasipoisson(link = log),offset = log(non_na_count))

    pm=summary(frail_mod)$p.table
    pm<-as.data.frame(pm,keep.rownames = TRUE)[] # Format as datatable
    table_pm=pm[2:4,]
    
    # get upper and lower bound
    table_pm$lower = table_pm$Estimate - qnorm(0.975) * table_pm$`Std. Error`
    table_pm$upper = table_pm$Estimate + qnorm(0.975) * table_pm$`Std. Error`
    table_pm = table_pm[,c(1,5:6) ]

    # PM row
    table_pm$Estimate_exp <- (exp(table_pm$Estimate*iqrs)-1)*100
    table_pm$lower_exp <- (exp(table_pm$lower*iqrs)-1)*100
    table_pm$upper_exp <- (exp(table_pm$upper*iqrs)-1)*100
    
  # Round values 
  table_pm$Estimate_exp <- round(table_pm$Estimate_exp, 3)
  table_pm$lower_exp <- round(table_pm$lower_exp, 3)
  table_pm$upper_exp <- round(table_pm$upper_exp, 3)
  table_pm$complete <- paste(table_pm$Estimate_exp, "(", table_pm$lower_exp, ";", table_pm$upper_exp, ")")
  table_pm[, 7, drop = FALSE]
```

# Model 2: PTSD analysis
```{r}
#Outcome model
ptsd_mod <- bam(as.numeric(PTSD)~ pm +temp_winter+ Mean_GVI_300 +age+s(year,bs='ps') + as.factor(Gender) + as.factor(Race4) + as.factor(edu4)+ pct_black + MEDHH + pct_poverty+pct_NoHS+as.factor(work2)+ as.factor(smoke2)+s(RunningID,bs='re'),family = quasipoisson(link = log), data=df_trimmed2)

    pm=summary(ptsd_mod)$p.table
    pm<-as.data.frame(pm,keep.rownames = TRUE)[] # Format as datatable
    table_pm=pm[2:4,]
    
    # get upper and lower bound
    table_pm$lower = table_pm$Estimate - qnorm(0.975) * table_pm$`Std. Error`
    table_pm$upper = table_pm$Estimate + qnorm(0.975) * table_pm$`Std. Error`
    table_pm = table_pm[,c(1,5:6) ]

    # PM row
    table_pm$Estimate_exp <- (exp(table_pm$Estimate*iqrs)-1)*100
    table_pm$lower_exp <- (exp(table_pm$lower*iqrs)-1)*100
    table_pm$upper_exp <- (exp(table_pm$upper*iqrs)-1)*100
    
  # Round values 
  table_pm$Estimate_exp <- round(table_pm$Estimate_exp, 3)
  table_pm$lower_exp <- round(table_pm$lower_exp, 3)
  table_pm$upper_exp <- round(table_pm$upper_exp, 3)
  table_pm$complete <- paste(table_pm$Estimate_exp, "(", table_pm$lower_exp, ";", table_pm$upper_exp, ")")
  table_pm[, 7, drop = FALSE]

```

# Model 3 - Depression outcome 0-9
```{r}
PHQ9_mod <- bam(as.numeric(depression)~ pm +temp_winter+ Mean_GVI_300 +age+s(year,bs='ps') + as.factor(Gender) + as.factor(Race4) + as.factor(edu4)+ pct_black + MEDHH + pct_poverty+pct_NoHS+as.factor(work2)+ as.factor(smoke2)+s(RunningID,bs='re'),family = quasipoisson(link = log),data=df_trimmed2) 

    pm=summary(PHQ9_mod)$p.table
    pm<-as.data.frame(pm,keep.rownames = TRUE)[] # Format as datatable
    table_pm=pm[2:4,]
    
    table_pm$lower = table_pm$Estimate - qnorm(0.975) * table_pm$`Std. Error`
    table_pm$upper = table_pm$Estimate + qnorm(0.975) * table_pm$`Std. Error`
    table_pm = table_pm[,c(1,5:6) ]

    # PM row
    table_pm$Estimate_exp <- (exp(table_pm$Estimate*iqrs)-1)*100
    table_pm$lower_exp <- (exp(table_pm$lower*iqrs)-1)*100
    table_pm$upper_exp <- (exp(table_pm$upper*iqrs)-1)*100
    
  # Round values 
  table_pm$Estimate_exp <- round(table_pm$Estimate_exp, 3)
  table_pm$lower_exp <- round(table_pm$lower_exp, 3)
  table_pm$upper_exp <- round(table_pm$upper_exp, 3)
  table_pm$complete <- paste(table_pm$Estimate_exp, "(", table_pm$lower_exp, ";", table_pm$upper_exp, ")")
  table_pm[, 7, drop = FALSE]

```

