---
title: "Untitled"
output: html_document
date: "2026-07-24"
---

This code represents the original analytic workflow used in the study. Since the original data cannot be shared due to data-use and confidentiality restrictions, we created a simulated dataset to demonstrate code reproducibility and analytic procedures. As a result, the outputs generated from the simulated data differ from the results presented in the manuscript.

The simulation data is located in the zip file attached to this code

```{r include=FALSE}
# install required pacakges
install.packages(plyr)
install.packages(dplyr)
install.packages(mgcv)
install.packages(table1)
install.packages(tableone)
install.packages(qgcomp)
install.packages(qgcompint)
install.packages(lubridate)
install.packages(ggplot2)
install.packages(tidyr)
install.packages(tidyverse)
install.packages(sf)
install.packages(patchwork)
```


# run libraries
```{r include=FALSE}
library(plyr)
library(dplyr)
library(mgcv)
library(table1)
library(tableone)
library(qgcomp)
library(qgcompint)
library(lubridate)
library(ggplot2)
library(tidyr)
library(tidyverse)
library(sf)
library(patchwork)
```


# Set working direcory
```{r}
wd = "C:/......"

```

# Import data

List of variables:
# year - monitoring year
# Gender- Male/Female
# Race4- Race in 4 categories
# edu4 -  Education level in 4 categories
# work2 - Occupation on 9/11
# exp911_imp2 - WTC exposure index (High/Very High=1/ Low/Intermediate=0)
# arr_9_11_13 - Arrival date (Arrived on 9/11-9/14 =1/ Arrived after 9/14 =0)
# horror - Witnessing horror (yes=1/ No=0)
# smoke2 -  Smoking status
# age - Age at monitoring visit
# non_na_count - Number of counted deficits
# pct_black - % Black population 
# pct_poverty - % under the poverty line
# pct_NoHS - % with no high school education
# MEDHH - Median household income 
# pm - PM2.5 (µg m-3)
# mean_temp - Mean temperature (° C)
# Mean_GVI_300 - Green View Index inside 300m buffer
# Mean_GVI_500 - Green View Index inside 500m buffer
# frailty.sum - Frailty index is calculated using 30 items that are collected during each monitoring visit, including deficits such as diabetes, fair/poor self-rated health, difficulty climbing several flights of stairs, abnormal heart examination, and self-reported mental health conditions. Each item is coded as a binary value where a response of Fair/Poor is considered as a “deficit” (and assigned a score of 1), and a response of Excellent/Very Good/Good is assigned a score of 0 (i.e., not a deficit). 
# PTSD - PTSD score based on the DSM-IV measurement 
# depression - Depression score based on PHQ9.

```{r}
data = read.csv(paste0(wd, '/Sim_Data.csv'))
head(data)
```

# Descriptive prep function: used to create Table 1
```{r}
# change decimal point in table
my.render.cont <- function(x) {
    with(stats.default(x), 
         c("",
          "Mean (SD)" = sprintf("%s (%s)",
                                round_pad(MEAN, 2),
                                round_pad(SD, 2))))
}

my.render.cat <- function(x) { 
  sub('.', '.', c("", sapply(stats.default(x), 
  function(y) with(y, sprintf("%d (%0.2f %%)", FREQ, PCT)))), fixed = TRUE) 
}

# 1. Define the missing value renderer to match your style
my.render.miss <- function(x, ...) {
  with(stats.default(x), 
       c("", "Missing" = sprintf("%d (%0.2f %%)", NMISS, PCTMISS)))
}
```

## Table 1. Population characteristics at cohort entry (N = 18,734), 2003-2023
```{r}
m2 <-data %>% group_by(RunningID) %>% filter(row_number()==1)

m2$outcome = 'Descriptive Statistics'

table1(~ age + as.factor(Race4) + as.factor(edu4) + as.factor(Gender) + 
           as.factor(smoke2) + as.factor(work2) + MEDHH + pct_black + 
           pct_poverty + pct_NoHS | outcome, 
       data = m2,
       overall = FALSE,
       render.continuous = my.render.cont, 
       render.categorical = my.render.cat,
       render.missing = render.missing.default)
```

# Table 2. Summary statistics of annual exposures and modifiers
```{r}
# Create mean continuous variables 
tab <- data %>%
    group_by(RunningID) %>%
    summarise(
        pm = mean(pm, na.rm = TRUE),
        mean_temp = mean(mean_temp, na.rm = TRUE),
        Mean_GVI_300 = mean(Mean_GVI_300, na.rm = TRUE),
        # categorical variables just take first (they don't change anyway)
        exp911_imp2 = first(exp911_imp2),
        arr_9_11_13 = first(arr_9_11_13),
        horror = first(horror)
    )

# Define variables
listVars <- c('pm','mean_temp','Mean_GVI_300','exp911_imp2','arr_9_11_13','horror')
catVars  <- c('exp911_imp2','arr_9_11_13','horror')

# Create Table 2
table2 <- CreateTableOne(vars = listVars,
                         data = tab,
                         factorVars = catVars,
                         includeNA = TRUE)

print(table2)
```
